---
title: "AIC and BIC for TT P-splines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AIC and BIC for TT P-splines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# What this vignette is

`TTPsplines` reports a **joint linearized EDF** after a fit. This note shows
how to turn that EDF into **AIC / BIC**-style scores for fixed-\(\lambda\)
comparisons (e.g. candidate ranks or anisotropic \(\lambda\)).

There is **no** `AIC()` / `BIC()` S3 method yet: compute the criteria from
`fit$deviance` and `fit$edf` as below. Rank selection for prediction still
belongs to `tt_rank_select()` (CV + 1-SE); see
`vignette("rank-selection")`.

```{r}
library(TTPsplines)
```

# Complexity layers (again)

| Quantity | Meaning | Use for AIC/BIC? |
|----------|---------|------------------|
| `npar_tt` | packed TT core entries | **No** |
| `npar_tt_intrinsic` | packed size minus gauge dim | **No** |
| `edf` | joint linearized effective df | **Yes** |

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

The EDF is
\[
\widehat{\mathrm{edf}}
=\operatorname{tr}\!\big[(J^\top J+P_{\boldsymbol\lambda})^{-1}J^\top J\big],
\]
with stacked conditional Jacobian \(J\) of the TT map at convergence and
block-diagonal core penalty \(P_{\boldsymbol\lambda}\). Gauge directions
contribute little; do **not** replace \(\widehat{\mathrm{edf}}\) by
`npar_tt`.

Enable it with `tt_control(compute_edf = TRUE)` (default in many demos;
skipped automatically when packed TT size exceeds `edf_max_npar`).

# Which number from `summary()`?

From a typical `summary(fit)` block, take:

```text
EDF (joint linearized): 54.64
```

Ignore `EDF / stored parameters` and `EDF / intrinsic TT dim.` — those are
diagnostic ratios only. The intercept is **not** inside the joint EDF, so
penalties below use \(\widehat{\mathrm{edf}}+1\).

# Working formulas

Helpers used in this vignette (not exported package API yet):

```{r helpers}
`%||%` <- function(a, b) if (!is.null(a)) a else b

tt_ic <- function(fit, criterion = c("AIC", "BIC")) {
  criterion <- match.arg(criterion)
  edf <- fit$edf
  if (!is.finite(edf)) {
    stop("fit$edf is missing; refit with tt_control(compute_edf = TRUE).")
  }
  n <- as.integer(fit$n %||% length(fitted(fit)))
  df <- edf + 1                       # + intercept
  pen <- if (identical(criterion, "AIC")) 2 else log(n)
  key <- fit$family_key %||% fit$family$family

  if (identical(key, "gaussian")) {
    # Gaussian: n log(RSS/n) + pen * df  (RSS stored in fit$deviance)
    return(n * log(fit$deviance / n) + pen * df)
  }
  # Poisson / Bernoulli: deviance + pen * df
  fit$deviance + pen * df
}
```

**Gaussian** (same as the lab coordinate-AIC notes):
\[
\mathrm{AIC}=n\log(\mathrm{RSS}/n)+2(\widehat{\mathrm{edf}}+1),\qquad
\mathrm{BIC}=n\log(\mathrm{RSS}/n)+\log(n)\,(\widehat{\mathrm{edf}}+1).
\]
Here `fit$deviance` is **RSS**.

**Poisson / Bernoulli:**
\[
\mathrm{AIC}=D+2(\widehat{\mathrm{edf}}+1),\qquad
\mathrm{BIC}=D+\log(n)\,(\widehat{\mathrm{edf}}+1),
\]
with \(D=\) GLM deviance (`fit$deviance` / `deviance(fit)`).

**Honesty:** \(\widehat{\mathrm{edf}}\) is for the **linearized** TT map
(PIRLS working weights for GLMs). This is a working IC, not an exact
`mgcv`-style REML score.

# Gaussian example

```{r gauss}
data(ishigami)
idx <- seq_len(400)
X <- as.matrix(ishigami[idx, c("x1", "x2", "x3")])
y <- ishigami$y[idx]

fit_g <- ttpspline(
  y, X,
  family = gaussian(),
  rank = 2, k = 8, lambda = 1,
  control = tt_control(max_sweeps = 8, backend = "R", compute_edf = TRUE)
)

c(
  RSS = fit_g$deviance,
  edf = fit_g$edf,
  npar_tt = fit_g$npar_tt,
  AIC = tt_ic(fit_g, "AIC"),
  BIC = tt_ic(fit_g, "BIC")
)
```

Compare two ranks on the **same** data / \(\lambda\) (in-sample IC, not CV):

```{r gauss-ranks}
fits <- lapply(1:3, function(r) {
  ttpspline(
    y, X, family = gaussian(), rank = r, k = 8, lambda = 1,
    control = tt_control(max_sweeps = 6, backend = "R", compute_edf = TRUE)
  )
})
data.frame(
  rank = 1:3,
  edf = vapply(fits, function(f) f$edf, numeric(1)),
  RSS = vapply(fits, function(f) f$deviance, numeric(1)),
  AIC = vapply(fits, function(f) tt_ic(f, "AIC"), numeric(1)),
  BIC = vapply(fits, function(f) tt_ic(f, "BIC"), numeric(1))
)
```

# Poisson example

```{r poisson}
set.seed(1)
n <- 500
Xp <- matrix(runif(n * 2), n, 2)
eta <- 0.5 + sin(2 * pi * Xp[, 1]) * cos(2 * pi * Xp[, 2])
yp <- rpois(n, exp(eta))

fit_p <- ttpspline(
  yp, Xp,
  family = poisson(),
  rank = 2, k = 10, lambda = c(10, 1),
  control = tt_control(
    pirls_maxit = 15, max_sweeps = 6,
    backend = "R", compute_edf = TRUE
  )
)

c(
  deviance = deviance(fit_p),
  edf = fit_p$edf,
  AIC = tt_ic(fit_p, "AIC"),
  BIC = tt_ic(fit_p, "BIC")
)
summary(fit_p)
```

# When to use AIC/BIC vs CV

| Goal | Prefer |
|------|--------|
| Predictive rank \(r\) | `tt_rank_select()` (CV + 1-SE) |
| In-sample \(\lambda\) / model comparison with an EDF penalty | AIC / BIC as above |
| Automatic anisotropic \(\lambda\) inside ALS | `vignette("cgcv")` now; coordinate AIC/BIC / TT-cFS later (Paper 2) |

AIC/BIC here are **conditional on** the chosen TT rank and fixed
\(\boldsymbol\lambda\). They do not fold in uncertainty from rank selection
or \(\lambda\) search.

# What this is not

- Not `AIC(fit)` / `BIC(fit)` methods (not implemented yet).
- Not REML / cREML.
- Not a substitute for hold-out or CV when the goal is prediction.
- Not a claim that linearized TT EDF equals dense-tensor EDF.

# See also

- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE for \(r\)
- `vignette("cgcv", package = "TTPsplines")` — fixed vs cGCV for \(\lambda\)
- `vignette("generalized", package = "TTPsplines")` — Poisson / Bernoulli
- `tt_complexity()`, `deviance()`, `summary()`

# References

Hastie, T. and Tibshirani, R. (1990).
*Generalized Additive Models*. Chapman & Hall.

Wood, S. N. (2017).
*Generalized Additive Models: An Introduction with R* (2nd ed.).
CRC Press.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
