---
title: "Array mode: product grids with array = TRUE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Array mode: product grids with array = TRUE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# What this vignette is for

Most of `TTPsplines` is about **scattered** covariates: an \(n\times d\) matrix
`X` and a length-\(n\) response `y`. That is the default.

Sometimes the response already lives on a **full product grid** — every
combination of margin coordinates is observed once (or with a known weight).
Classic examples: age × year mortality tables, temperature × lag surfaces on a
regular mesh, or a simulated \(n_1\times n_2\times n_3\) array.

For that layout, call

```r
ttps(Y, axes = list(...), array = TRUE, ...)
```

You pass the **array** `Y` and the **margin axes**, not a flattened
`expand.grid` design. Statistically you still fit the same TT P-spline; the
difference is **how the ALS Gram / RHS are formed** (Kronecker / mode
contractions instead of an \(n\times q_k\) design matrix).

```{r}
library(TTPsplines)
```

# Three related tools (do not confuse them)

| Tool | Data layout | Coefficient storage | Typical use |
|------|-------------|---------------------|-------------|
| `ttps(..., array = FALSE)` | Scattered \(X\) | TT cores | Irregular designs (default) |
| `ttps(..., array = TRUE)` | Full product array `Y` | TT cores | Dense grids; faster Gram/RHS |
| `glam_fit_poisson()` | Full product array | **Dense** \(\Theta\) | Classical Currie–Durbán–Eilers GLAM |

So:

- **GLAM** (`vignette("glam-poisson")`, `vignette("glam-vs-tt")`) = array
  *data* + dense coefficient array.
- **`array = TRUE`** = array *data* + **TT** coefficients (this vignette).
- **Scattered `ttps()`** = no grid required + TT coefficients
  (`vignette("getting-started")`).

\[
\boxed{\texttt{array = TRUE}\;\neq\;\text{GLAM}}
\qquad
\boxed{\texttt{array = TRUE}\;=\;\text{same TT model, grid algebra}}
\]

# Building the inputs

You need two things:

1. A numeric array `Y` of dimension \(n_1\times\cdots\times n_d\) (\(d\ge 2\)).
2. A list `axes` of length \(d\): `axes[[j]]` is the length-\(n_j\) coordinate
   vector for margin \(j\).

## Memory layout (important)

R stores arrays with **the first dimension varying fastest**. That is the same
order as

```r
as.numeric(Y)                    # length prod(dim(Y))
expand.grid(axes)                # row i matches as.numeric(Y)[i]
```

So if you build `Y` from a vector `y` and an `expand.grid` design, use

```r
Y <- array(y, dim = c(n1, n2, ..., nd))
```

not a transpose of that layout. Getting this wrong silently misaligns cells
with axes.

## A small 2D example

```{r build-2d}
set.seed(1)
n1 <- 20L
n2 <- 16L
axes <- list(
  x1 = seq(0, 1, length.out = n1),
  x2 = seq(0, 1, length.out = n2)
)

# Truth on the product grid (outer = all pairs)
mu <- outer(axes$x1, axes$x2, function(a, b) {
  sin(2 * pi * a) * cos(pi * b) + 0.3 * a
})
Y <- mu + array(rnorm(n1 * n2, sd = 0.15), dim = c(n1, n2))

dim(Y)
str(axes, vec.len = 2)
```

If you omit `axes`, the package assumes a unit-interval grid
`seq(0, 1, length.out = n_j)` on each margin. Prefer passing real axes when
coordinates matter (age, year, temperature, …).

# First fit

```{r fit-array}
fit <- ttps(
  Y,
  axes = axes,
  array = TRUE,
  rank = 3L,
  k = 8L,
  lambda = 1,                 # fixed isotropic penalty for a clear demo
  family = gaussian(),
  control = tt_control(
    max_sweeps = 8L,
    compute_edf = FALSE,
    seed = 1L,
    trace = FALSE
  )
)

fit$backend          # should be "R-array"
c(npar_tt = fit$npar_tt, npar_dense = fit$npar_dense)
head(fit$lambda)
```

`backend = "R-array"` means the ALS sweeps used the array Gram/RHS path.
Fitted values are still a **length-\(n\) vector** in dim1-fastest order — the
same order as `as.numeric(Y)`. To reshape them back to the grid:

```{r reshape}
mu_hat <- array(fitted(fit), dim = dim(Y))
rmse <- sqrt(mean((mu_hat - mu)^2))
rmse
```

```{r plot-slice, fig.cap = "True vs fitted surface (2D grid)."}
zlim <- range(mu, mu_hat)
opar <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
image(axes$x1, axes$x2, mu, zlim = zlim,
      xlab = "x1", ylab = "x2", main = "Truth")
image(axes$x1, axes$x2, mu_hat, zlim = zlim,
      xlab = "x1", ylab = "x2", main = "Fitted (array = TRUE)")
par(opar)
```

Predict at new (possibly non-grid) points with the usual API:

```{r predict}
X_new <- cbind(x1 = c(0.2, 0.5, 0.8), x2 = c(0.3, 0.5, 0.7))
predict(fit, X_new, type = "link")
```

# Same model as scattered — different algebra

If you flatten the same grid into `(y, X)`, `array = FALSE` estimates the
**same** TT P-spline. With a shared random seed and one ALS sweep, fitted
values agree to near machine precision (the package tests this).

```{r parity}
idx <- expand.grid(i1 = seq_len(n1), i2 = seq_len(n2), KEEP.OUT.ATTRS = FALSE)
X_sc <- cbind(x1 = axes$x1[idx$i1], x2 = axes$x2[idx$i2])
y_sc <- as.numeric(Y)

ctrl1 <- tt_control(
  max_sweeps = 1L, compute_edf = FALSE, seed = 1L, trace = FALSE
)
fit_sc <- ttps(y_sc, X_sc, rank = 3L, k = 8L, lambda = 1,
               control = ctrl1)
fit_ar <- ttps(Y, axes = axes, array = TRUE, rank = 3L, k = 8L, lambda = 1,
               control = ctrl1)

max_abs <- max(abs(fitted(fit_ar) - fitted(fit_sc)))
c(max_abs_diff = max_abs, backend_array = fit_ar$backend,
  backend_scattered = fit_sc$backend)
```

Use `array = TRUE` when the grid is **complete** (or almost complete with
`weights`; see below) and you want the cheaper contractions. Use scattered
when points are irregular, incomplete, or already stored as rows.

# Choosing \(\lambda\) on a grid (cGCV)

Fixed \(\lambda\) is fine for demos. For anisotropic directional penalties,
the default selector `"cGCV"` works in array mode (same conditional GCV idea
as scattered; see `vignette("cgcv")`).

```{r cgcv}
fit_cgcv <- ttps(
  Y, axes = axes, array = TRUE,
  rank = 3L, k = 8L, lambda = "cGCV",
  control = tt_control(
    max_sweeps = 6L, compute_edf = FALSE, seed = 1L,
    cgcv_update = "sequential", trace = FALSE
  )
)
round(fit_cgcv$lambda, 4)
fit_cgcv$lambda_method
```

**Not** available with `array = TRUE` in this version: `lambda = "CV"` and
`lambda = "gGCV"`. Use fixed \(\lambda\) or `"cGCV"`.

# A 3D walkthrough (still small)

```{r build-3d}
set.seed(7)
ng <- c(10L, 9L, 8L)
axes3 <- list(
  x1 = seq(0, 1, length.out = ng[1]),
  x2 = seq(0, 1, length.out = ng[2]),
  x3 = seq(0, 1, length.out = ng[3])
)
# Separable-ish truth: low TT rank is enough
g <- expand.grid(axes3, KEEP.OUT.ATTRS = FALSE)
mu3 <- with(g, sin(2 * pi * x1) * cos(2 * pi * x2) + 0.5 * x3)
Y3 <- array(mu3 + rnorm(prod(ng), sd = 0.2), dim = ng)

fit3 <- ttps(
  Y3, axes = axes3, array = TRUE,
  rank = 2L, k = 6L, lambda = 1,
  control = tt_control(max_sweeps = 6L, compute_edf = FALSE, seed = 1L)
)
c(
  n = prod(ng),
  npar_tt = fit3$npar_tt,
  npar_dense = fit3$npar_dense,
  compression = round(fit3$compression_ratio, 2),
  rmse = sqrt(mean((fitted(fit3) - mu3)^2))
)
```

Compression here is about **coefficients** (\(N_{\mathrm{TT}}\) vs
\(k^d\)), not about storing `Y`. The response array still has
\(\prod_j n_j\) cells.

# Poisson grids (short note)

Array mode also accepts Poisson responses (PIRLS-ALS / weighted row-tensor
Gram). For a **classical exposure-weighted mortality table** with dense
\(\Theta\), prefer `glam_fit_poisson()` (`vignette("glam-poisson")`).

```{r poisson, fig.cap = "Poisson array fit: log mean vs truth (diagonal = perfect)."}
set.seed(21)
eta <- outer(axes$x1, axes$x2, function(a, b) {
  0.4 + sin(pi * a) + 0.5 * cos(pi * b)
})
Yp <- array(rpois(n1 * n2, exp(eta)), dim = c(n1, n2))

fit_p <- ttps(
  Yp, axes = axes, array = TRUE,
  family = poisson(),
  rank = 2L, k = 6L, lambda = 1,
  control = tt_control(
    max_sweeps = 4L, pirls_maxit = 8L,
    compute_edf = FALSE, seed = 1L, trace = FALSE
  )
)
plot(as.numeric(eta), log(pmax(fitted(fit_p), 1e-8)),
     xlab = "True log mean", ylab = "Fitted log mean",
     pch = 16, cex = 0.5)
abline(0, 1, col = 2, lwd = 2)
```

Empty or unreliable cells can be down-weighted with `weights`
(zeros drop them from the likelihood). Incomplete *designs* that are not
product-shaped should stay in scattered mode.

# Current restrictions (checklist)

Keep this list in mind before teaching `array = TRUE` as a drop-in for every
grid problem:

- `y` must be a **d-way array** (\(d\ge 2\)); vectors error.
- `axes` length and lengths of each vector must match `dim(y)`.
- No `linear=` / `smooth=` terms.
- `null_space = "joint"` only (not `"profiled"`).
- No `lambda = "CV"` / `"gGCV"`.
- Gaussian unweighted grids use the Kronecker Gram path; GLM / weighted grids
  use the weighted array Gram (still without forming \(X_k\)).
- Rank \(r\) is still **user-chosen** (see `vignette("rank-selection")`).

# When to prefer which path

- **Scattered points** → `ttps(y, X)` (default).
- **Full rectangular Gaussian / Poisson grid, TT coefficients** →
  `ttps(Y, axes = ..., array = TRUE)`.
- **Dense GLAM Poisson with exposure, classical baseline** →
  `glam_fit_poisson()` then optionally compare to TT
  (`vignette("glam-vs-tt")`).
- **Missing / irregular locations** → flatten only the observed cells and use
  scattered `ttps()`, or mask with `weights` if the array shell is still
  product-shaped.

# See also

- `vignette("getting-started")` — scattered workflow
- `vignette("cgcv")` — \(\lambda\) selectors
- `vignette("glam-poisson")` — Currie–Durbán–Eilers Poisson GLAM
- `vignette("glam-vs-tt")` — dense full-tensor vs TT storage
- `vignette("scalability")` — \(k^d\) vs \(O(dkr^2)\)

# References

Currie, I. D., Durbán, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280. <https://doi.org/10.1111/j.1467-9868.2006.00543.x>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>
