---
title: "Choosing λ with conditional GCV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing λ with conditional GCV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# What this vignette is

Directional smoothing \(\boldsymbol\lambda=(\lambda_1,\ldots,\lambda_d)\) is
**separate** from the TT rank \(r\):

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

This note covers **fixed** \(\lambda\) vs **conditional GCV (cGCV)** inside
`ttps()`. Rank selection is `vignette("rank-selection")`; in-sample
AIC/BIC is `vignette("aic-bic")`.

```{r}
library(TTPsplines)
data(sobol_g)
idx <- seq_len(400)
X <- as.matrix(sobol_g[idx, paste0("x", 1:4)])
y <- sobol_g$y[idx]
```

# Fixed \(\lambda\)

Pass a scalar (isotropic) or length-\(d\) anisotropic vector:

```{r fixed}
fit_fix <- ttps(
  y, X,
  family = gaussian(),
  rank = 2, k = 6,
  lambda = c(0.1, 1, 10, 100),
  control = tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE)
)
fit_fix$lambda_method
fit_fix$lambda
```

# Conditional GCV

`lambda = "cGCV"` selects each \(\lambda_k\) by **conditional** GCV while
updating TT cores (ALS / PIRLS), not by a full \(d\)-dimensional grid over
the joint map.

**Default dynamics** (Chicago-validated under the exact global penalty):
`cgcv_update = "outer_simultaneous"` — fit all cores at fixed
\(\boldsymbol\lambda\), freeze, Jacobi-propose every margin, then apply
log-damping (`cgcv_damping = 0.25`) and a trust region
(`cgcv_max_log10_step = 1`). Legacy `"sequential"` Gauss–Seidel can cascade
to \(\lambda_{\max}\) via immediate cross-margin \(P_{k,-k}\) feedback; keep
it for diagnostics only. See `fit$cgcv` and `tt_cgcv_frozen_curves()`.

```{r cgcv}
fit <- ttps(
  y, X,
  family = gaussian(),
  rank = 2, k = 6,
  lambda = "cGCV",
  control = tt_control(
    max_sweeps = 8,
    backend = "R",
    lambda_start = 1,
    lambda_bounds = c(1e-4, 1e4),
    warn_lambda_boundary = TRUE,
    compute_edf = FALSE,
    outer_maxit = 6
    # defaults: cgcv_update = "outer_simultaneous",
    # cgcv_damping = 0.25, cgcv_max_log10_step = 1
  )
)
cgcv_upd <- if (!is.null(fit$cgcv$update)) fit$cgcv$update else NA_character_
c(
  method = fit$lambda_method,
  update = cgcv_upd,
  lambda = paste(sprintf("%.4g", fit$lambda), collapse = ", "),
  boundary = paste(fit$lambda_boundary, collapse = ", ")
)
```

Useful control knobs:

| Control | Role |
|---------|------|
| `cgcv_update` | `"outer_simultaneous"` (default) or `"sequential"` |
| `cgcv_damping` | Log-mix \(\rho\) toward the raw proposal |
| `cgcv_max_log10_step` | Trust region on \(|\Delta\log_{10}\lambda|\) |
| `lambda_start` | Initial \(\lambda\) (recycled to length \(d\)) |
| `lambda_bounds` | Log-scale search interval \([\lambda_{\min},\lambda_{\max}]\) |
| `tol_lambda` | Relative log-\(\lambda\) change for outer convergence |
| `warn_lambda_boundary` | Soft warning when a margin hits a bound |

Inspect `fit$lambda`, `fit$lambda_boundary`, `fit$cgcv`, and
`fit$lambda_bounds` after fitting. Bound hits under `"sequential"` are often
update-dynamics artefacts — switch to outer simultaneous before interpreting
them as scientific nulls.

# Multi-start diagnostics (nonconvex ALS)

`ttps()` uses a **single** random TT initialization. Because the core
problem is nonconvex, a \(\lambda\) on the search bound can mean either
(i)~a statistical effect (near-null margin, low \(r\), conditional cGCV) or
(ii)~a bad basin. Use `ttps_multistart()` to tell them apart:

```{r multistart, fig.show='hide'}
ms <- ttps_multistart(
  y, X[, 1:2],
  rank = 2, k = 5,
  lambda = "cGCV",
  n_starts = 4,
  seed = 1,
  control = tt_control(
    max_sweeps = 5, backend = "R", compute_edf = FALSE,
    lambda_bounds = c(1e-4, 1e4)
  )
)
ms
ms$starts[, c("start", "converged", "objective", "lambda_at_boundary",
              "lambda_boundary")]
ms$frac_boundary
# Retain the selected fit:
fit_best <- ms$best
```

- **Stable** boundary hits across starts (\(\texttt{frac\_boundary}\) near 1)
  → interpret as margin / null space / rank, not a lucky seed.
- **Unstable** hits → raise `n_starts`, try another optimizer, or warm-start.

For multi-start inside **rank CV**, prefer
`tt_rank_select(..., n_starts = ...)` (already fold-aware).

# Anisotropy on Sobol-\(g\)

Earlier coordinates of Sobol-\(g\) matter more; cGCV often pushes
\(\lambda_k\) apart:

```{r anisotropy}
fit_iso <- ttps(
  y, X, rank = 2, k = 6, lambda = 1,
  control = tt_control(max_sweeps = 6, backend = "R", compute_edf = FALSE)
)
data.frame(
  margin = seq_len(ncol(X)),
  lambda_fixed_iso = fit_iso$lambda,
  lambda_cGCV = fit$lambda,
  boundary = fit$lambda_boundary
)
```

# With GLMs

The same `lambda = "cGCV"` works for Poisson (PIRLS-ALS). Bernoulli /
L-BFGS paths also accept `"cGCV"`, but separation / bound hits are more
common — check `lambda_boundary`.

```{r poisson, warning=TRUE}
set.seed(2)
n <- 400
Xp <- matrix(runif(n * 2), n, 2)
eta <- 0.4 + sin(2 * pi * Xp[, 1]) * cos(2 * pi * Xp[, 2])
yp <- rpois(n, exp(eta))
fit_p <- ttps(
  yp, Xp,
  family = poisson(),
  rank = 2, k = 8,
  lambda = "cGCV",
  control = tt_control(
    pirls_maxit = 12, max_sweeps = 4,
    backend = "R", compute_edf = FALSE
  )
)
fit_p$lambda
fit_p$lambda_boundary
```

# cGCV inside rank CV

When selecting \(r\) with `tt_rank_select(..., lambda = "cGCV")`, each
training fold runs its **own** cGCV; held-out rows never enter the λ
search. That is honest but slower than fixed \(\lambda\).

```{r rank-cgcv, eval = FALSE}
sel <- tt_rank_select(
  y, X, ranks = 1:3, k = 5, lambda = "cGCV",
  folds = 3, rule = "1se", seed = 1,
  control = tt_control(max_sweeps = 4, compute_edf = FALSE)
)
```

# What this is not

- Not REML / cREML (reserved: `lambda = "cREML"` errors for now).
- Not TT-cFS / Fellner–Schall inside ALS (`lambda = "cFS"` not implemented).
- Not coordinate AIC/BIC λ search (lab prototype; IC formulas in
  `vignette("aic-bic")`).
- Not a substitute for choosing TT rank \(r\).

# See also

- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE for \(r\)
- `vignette("aic-bic", package = "TTPsplines")` — AIC / BIC from EDF
- `vignette("getting-started", package = "TTPsplines")`
- `vignette("generalized", package = "TTPsplines")`

# References

Craven, P. and Wahba, G. (1979).
Smoothing noisy data with spline functions.
*Numerische Mathematik*, **31**, 377–403.

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
