---
title: "Choosing λ (cGCV, CV, gGCV)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing λ (cGCV, CV, gGCV)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# What this vignette is

Directional smoothing \(\boldsymbol\lambda=(\lambda_1,\ldots,\lambda_d)\) is
**separate** from the TT rank \(r\):

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

This note covers \(\lambda\) selectors inside `ttps()`:

| Spec | Role |
|------|------|
| numeric / length-`d` | Fixed isotropic or anisotropic |
| `"cGCV"` (default) | Conditional / product GCV inside ALS / PIRLS |
| `"CV"` | K-fold CV of each \(\lambda_k\) (ALS / PIRLS) |
| `"gGCV"` | Global GCV / UBRE with a Monte Carlo GDF (Gaussian or Poisson; scattered or array; expensive) |

Rank selection is `vignette("rank-selection")`; in-sample AIC/BIC is
`vignette("aic-bic")`.

```{r}
library(TTPsplines)
data(sobol_g)
idx <- seq_len(400)
X <- as.matrix(sobol_g[idx, paste0("x", 1:4)])
y <- sobol_g$y[idx]
```

# Fixed \(\lambda\)

Pass a scalar (isotropic) or length-\(d\) anisotropic vector:

```{r fixed}
fit_fix <- ttps(
  y, X,
  family = gaussian(),
  rank = 2, k = 6,
  lambda = c(0.1, 1, 10, 100),
  control = tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE)
)
fit_fix$lambda_method
fit_fix$lambda
```

# Conditional GCV

`lambda = "cGCV"` selects each \(\lambda_k\) by **conditional** GCV while
updating TT cores (ALS / PIRLS), not by a full \(d\)-dimensional grid over
the joint map.

**Default dynamics** (exact global penalty):
`cgcv_update = "outer_simultaneous"` — fit all cores at fixed
\(\boldsymbol\lambda\), freeze, Jacobi-propose every margin, then apply
log-damping (`cgcv_damping = 0.25`) and a trust region
(`cgcv_max_log10_step = 1`). Legacy `"sequential"` Gauss–Seidel can cascade
to \(\lambda_{\max}\) via immediate cross-margin \(P_{k,-k}\) feedback; keep
it for diagnostics only. See `fit$cgcv` and `tt_cgcv_frozen_curves()`.

```{r cgcv}
fit <- ttps(
  y, X,
  family = gaussian(),
  rank = 2, k = 6,
  lambda = "cGCV",
  control = tt_control(
    max_sweeps = 8,
    backend = "R",
    lambda_start = 1,
    lambda_bounds = c(1e-4, 1e4),
    warn_lambda_boundary = TRUE,
    compute_edf = FALSE,
    outer_maxit = 6
    # defaults: cgcv_update = "outer_simultaneous",
    # cgcv_damping = 0.25, cgcv_max_log10_step = 1
  )
)
cgcv_upd <- if (!is.null(fit$cgcv$update)) fit$cgcv$update else NA_character_
c(
  method = fit$lambda_method,
  update = cgcv_upd,
  lambda = paste(sprintf("%.4g", fit$lambda), collapse = ", "),
  boundary = paste(fit$lambda_boundary, collapse = ", ")
)
```

Useful control knobs:

| Control | Role |
|---------|------|
| `cgcv_update` | `"outer_simultaneous"` (default) or `"sequential"` |
| `cgcv_damping` | Log-mix \(\rho\) toward the raw proposal |
| `cgcv_max_log10_step` | Trust region on \(|\Delta\log_{10}\lambda|\) |
| `lambda_start` | Initial \(\lambda\) (recycled to length \(d\)) |
| `lambda_bounds` | Log-scale search interval \([\lambda_{\min},\lambda_{\max}]\) |
| `tol_lambda` | Relative log-\(\lambda\) change for outer convergence |
| `warn_lambda_boundary` | Soft warning when a margin hits a bound |

Inspect `fit$lambda`, `fit$lambda_boundary`, `fit$cgcv`, and
`fit$lambda_bounds` after fitting. Bound hits under `"sequential"` are often
update-dynamics artefacts — switch to outer simultaneous before interpreting
them as scientific nulls.

# Multi-start diagnostics (nonconvex ALS)

`ttps()` uses a **single** random TT initialization. Use multi-start only
when you have a reason to doubt that start — not as the default path.

**When to run it**

- a margin hits `lambda_boundary` under cGCV;
- fits change materially if you change `seed`;
- low rank (\(r = 1,2\)) looks unstable.

**Practical workflow**

1. Fit once with `ttps(..., lambda = "cGCV")`.
2. If nothing looks suspicious, stop.
3. Otherwise call `ttps_multistart()` (typically `n_starts = 5`–`10`) and
   keep `ms$best`.
4. Read `ms$frac_boundary`: near 1 on margin \(k\) → statistical
   null / weak margin / low \(r\); unstable across starts → more starts or
   another optimizer.

```{r multistart, fig.show='hide'}
ms <- ttps_multistart(
  y, X[, 1:2],
  rank = 2, k = 5,
  lambda = "cGCV",
  n_starts = 4,
  seed = 1,
  control = tt_control(
    max_sweeps = 5, backend = "R", compute_edf = FALSE,
    lambda_bounds = c(1e-4, 1e4)
  )
)
ms
ms$starts[, c("start", "converged", "objective", "lambda_at_boundary",
              "lambda_boundary")]
ms$frac_boundary
# Retain the selected fit:
fit_best <- ms$best
```

- **Stable** boundary hits across starts (`frac_boundary` near 1)
  → interpret as margin / null space / rank, not a lucky seed.
- **Unstable** hits → raise `n_starts`, try another optimizer, or warm-start.

Cost scales as `n_starts` times one `ttps()` fit. For multi-start inside
**rank CV**, prefer `tt_rank_select(..., n_starts = ...)` (already
fold-aware) rather than wrapping `ttps_multistart()` yourself.

# Anisotropy on Sobol-\(g\)

Earlier coordinates of Sobol-\(g\) matter more; cGCV often pushes
\(\lambda_k\) apart:

```{r anisotropy}
fit_iso <- ttps(
  y, X, rank = 2, k = 6, lambda = 1,
  control = tt_control(max_sweeps = 6, backend = "R", compute_edf = FALSE)
)
data.frame(
  margin = seq_len(ncol(X)),
  lambda_fixed_iso = fit_iso$lambda,
  lambda_cGCV = fit$lambda,
  boundary = fit$lambda_boundary
)
```

# With GLMs

The same `lambda = "cGCV"` (and `"CV"` under PIRLS-ALS) works for Poisson.
Bernoulli / L-BFGS paths also accept `"cGCV"`, but separation / bound hits
are more common — check `lambda_boundary`. `"CV"` requires ALS / PIRLS-ALS
(not global L-BFGS). `"gGCV"` covers Gaussian and Poisson responses (UBRE
by default for Poisson; see below), not Bernoulli.

```{r poisson, warning=TRUE}
set.seed(2)
n <- 400
Xp <- matrix(runif(n * 2), n, 2)
eta <- 0.4 + sin(2 * pi * Xp[, 1]) * cos(2 * pi * Xp[, 2])
yp <- rpois(n, exp(eta))
fit_p <- ttps(
  yp, Xp,
  family = poisson(),
  rank = 2, k = 8,
  lambda = "cGCV",
  control = tt_control(
    pirls_maxit = 12, max_sweeps = 4,
    backend = "R", compute_edf = FALSE
  )
)
fit_p$lambda
fit_p$lambda_boundary
```

# cGCV inside rank CV

When selecting \(r\) with `tt_rank_select(..., lambda = "cGCV")`, each
training fold runs its **own** cGCV; held-out rows never enter the λ
search. That is honest but slower than fixed \(\lambda\).

```{r rank-cgcv, eval = FALSE}
sel <- tt_rank_select(
  y, X, ranks = 1:3, k = 5, lambda = "cGCV",
  folds = 3, rule = "1se", seed = 1,
  control = tt_control(max_sweeps = 4, compute_edf = FALSE)
)
```

# K-fold CV (`lambda = "CV"`)

`lambda = "CV"` runs a **k-fold** grid search for each core's \(\lambda_k\)
during ALS (Gaussian) or PIRLS (Poisson), then continues the fit at the
chosen values. Default behaviour (Rafa): tune on the **first** sweep only
(`cv_sweeps = 1`), then freeze \(\boldsymbol\lambda\).

Requires ALS / PIRLS-ALS (not global L-BFGS). Score is MSE on held-out
working residuals for the current core design.

```{r cv}
fit_cv <- ttps(
  y, X[, 1:2],
  family = gaussian(),
  rank = 1, k = 5,
  lambda = "CV",
  control = tt_control(
    max_sweeps = 6,
    backend = "R",
    compute_edf = FALSE,
    cv_folds = 3,
    cv_ngrid = 7,
    cv_rule = "min",
    cv_sweeps = 1,
    seed = 3,
    warn_lambda_boundary = FALSE
  )
)
c(
  method = fit_cv$lambda_method,
  lambda = paste(sprintf("%.4g", fit_cv$lambda), collapse = ", "),
  folds = fit_cv$cv$folds,
  rule = fit_cv$cv$rule,
  sweeps = fit_cv$cv$sweeps
)
head(fit_cv$cv$trace)
```

Useful knobs:

| Control | Role |
|---------|------|
| `cv_folds` | Number of folds (default `5`) |
| `cv_ngrid` | Log-grid size over `lambda_bounds` if `cv_grid` is `NULL` |
| `cv_grid` | Explicit positive grid (overrides `cv_ngrid`) |
| `cv_sweeps` | How many ALS/PIRLS iterations retune before freeze (`1`, larger, or `Inf`) |
| `cv_rule` | `"min"` (default) or `"1se"` (larger \(\lambda\), less under-smoothing) |

`"1se"` keeps the **largest** \(\lambda\) whose CV score is within one SE of
the minimum — useful when `"min"` under-smooths.

# Global TT-gGCV (`lambda = "gGCV"` / `tt_ggcv()`)

Product cGCV is the **default**. **gGCV** scores the whole
fixed-\(\boldsymbol\lambda\) fit with one global criterion and searches
\(\boldsymbol\lambda\) directly:

\[
\mathrm{GCV}(\boldsymbol\lambda) = \frac{n\,D}{(n - \mathrm{GDF})^2}
\ \text{(Gaussian)}, \qquad
\mathrm{UBRE}(\boldsymbol\lambda) = \frac{D + 2\,\mathrm{GDF}}{n} - 1
\ \text{(Poisson)},
\]

with \(D\) the deviance and GDF the trace of the derivative of the fitting
map \(y \mapsto \hat\mu(y; \boldsymbol\lambda)\). `criterion = "auto"` picks
GCV for Gaussian and UBRE (known scale 1) for Poisson, as `mgcv`'s
`"GCV.Cp"`. GDF is a Monte Carlo trace: non-negative Rademacher
perturbations against one shared reference fit (Poisson counts are never
clipped at zero), the same probes at every \(\boldsymbol\lambda\), and, at
each \(\boldsymbol\lambda\), one iteration budget B for the fit at \(y\),
the reference and all probes from one cold initialization, so that GDF is
the derivative of one smooth map. The budget is adaptive by default: a
pilot at \(y\) finds the iteration \(n\) after which the penalized
objective has settled (relative change at most `fit_tol` at two
consecutive iterations) under a cap (`max_sweeps` / `pirls_maxit`), and
\(B = \lceil 1.5\,n \rceil + 2\), because a budget that suits one
\(\boldsymbol\lambda\) need not suit another (slowly mixing low-rank ALS).
An evaluation whose probe contributions are non-finite or outlying (a
per-probe coefficient of variation above a limit that is 1 for large GDF
and larger for small GDF) is re-evaluated at 2B and 4B, and flagged
unstable if it stays so; unstable evaluations do not steer the search or
win.
gGCV works for Gaussian and Poisson responses,
on scattered rows (`tt_ggcv()`, offsets allowed, e.g. Berman–Turner
quadrature rows) and on product grids (`array = TRUE`, `tt_ggcv_array()`,
cell weights and offsets allowed). Each new \(\boldsymbol\lambda\) costs
`M + 2` fits for `M` probes, so gGCV is much slower than cGCV.

**Search.** An isotropic grid on the shared \(\log_{10}\lambda\), a few
halvings of the grid step around the best point and one parabolic step;
with smoothing groups (`groups`, e.g. space versus time), a coordinate
pattern search from the isotropic optimum follows. The best candidates and
the nearest smoother isotropic point are re-scored with `M_final` probes.
Scores are compared on the AIC scale (\(D / \phi + 2\,\mathrm{GDF}\) for
UBRE with scale \(\phi\), or \(n \log \mathrm{GCV}\)), which does not
depend on the units of the response, against `max(tol, 2 * SE)`, where SE is
the paired Monte Carlo standard error over the common probes. The winner is the most
regular final candidate that the minimum does not beat by this margin
(isotropic before grouped, then the largest \(\lambda\); `decision`), in
the spirit of the one-standard-error rule: on the Dette d = 8 example with
Berman–Turner rows the plain argmin sat on the grid boundary, where the
Monte Carlo GDF was 432 ± 334, against 145 ± 10 at the \(\lambda\) chosen
by thinning CV. The winner's GDF is then re-estimated at twice its
iteration budget; if it moves by more than 1%, the winner is rescored with
the less favourable of its two scores and kept if it still wins,
otherwise the next candidate by the same rule is checked
(`budget_checks`, `budget_verified`). The first winner is also kept when
no other converged candidate with a stable GDF is left to check.

Use the helper when you want the search diagnostics:

```{r ggcv, eval = FALSE}
opt <- tt_ggcv(
  y, X[, 1:2],
  rank = 2, k = 5,
  M_search = 4, M_final = 8
)
opt$lambda
opt$decision
opt$paired         # final candidates: qualifies, best, winner, budget_ok
head(opt$search)   # one row per computed evaluation (budget B, n_iter)
fit_g <- opt$fit

# GDF and scores at a given lambda (array data: tt_gdf_array())
tt_gdf(y, X[, 1:2], lambda = opt$lambda, rank = 2, k = 5, M = 16)$gdf
```

Through `ttps()` the same selection runs with the `ggcv_*` knobs of
`tt_control()`; the model is then refitted at the selected
\(\boldsymbol\lambda\) with your own arguments and control (so
`compute_edf`, `tol`, `init` and `optimizer` apply), and the selection
diagnostics are kept in `fit$ggcv`:

```{r ggcv-ttps, eval = FALSE}
fit_g <- ttps(
  y, X[, 1:2],
  rank = 2, k = 5,
  lambda = "gGCV",
  control = tt_control(
    max_sweeps = 20, compute_edf = TRUE,
    ggcv_groups = 1:2, ggcv_M_search = 4, ggcv_M_final = 8
  )
)
fit_g$lambda
fit_g$ggcv[c("criterion", "gdf", "gdf_se", "decision", "boundary")]
```

| Control | Role |
|---------|------|
| `ggcv_groups` | Margins sharing one \(\lambda\) (`NULL` = isotropic, `1:d` = one per margin) |
| `ggcv_n_grid`, `ggcv_n_refine` | Isotropic grid size and halvings (`9`, `2`) |
| `ggcv_n_global` | Sobol starts of the grouped stage (`NULL` = none) |
| `ggcv_n_final` | Candidates re-scored with `ggcv_M_final` probes (`3`) |
| `ggcv_M_search`, `ggcv_M_final` | Monte Carlo probes (`4`, `16`) |
| `ggcv_tol` | Minimum improvement of the comparison score, AIC scale (`1`) |
| `ggcv_criterion` | `"auto"`, `"gcv"` or `"ubre"` |
| `ggcv_budget` | `"adaptive"` (default) or `"fixed"` iteration budget |
| `ggcv_fit_tol` | Convergence tolerance of the adaptive budget: relative change of the penalized objective (`1e-7`) |
| `ggcv_max_sweeps`, `ggcv_pirls_maxit` | Iteration caps of the selection fits (`400`, `60`); the fixed budget with `"fixed"` |
| `ggcv_budget_check` | Re-estimate the selected GDF at twice its budget (`TRUE`) |
| `ggcv_budget_fallback` | Rescore a winner that fails that check with its less favourable score and choose again (`TRUE`) |
| `ggcv_n_cores` | Forked workers per batch of fits (`1`) |
| `ggcv_refit` | Refit at the selected \(\lambda\) with your arguments (`TRUE`) |

With `array = TRUE`, knobs named `ggcv_array_<knob>` (set with `$<-` on the
control object) take precedence over `ggcv_<knob>`. Warnings flag a selected
\(\lambda\) at the search boundary, a selection fit that reached the cap
without converging, a GDF that moves by more than 1% when the budget doubles
(with the margin by which a kept winner still wins, or the candidates the
selection moved past), and an unstable Monte Carlo GDF; raise
`ggcv_pirls_maxit` (Poisson) or `ggcv_max_sweeps` (Gaussian), lower
`ggcv_fit_tol`, or increase `ggcv_M_final`. Not yet supported:
`linear=` / `smooth=` terms and non-uniform weights on scattered rows.

# What this is not

- Not REML.
- Not a claim that `"CV"` or `"gGCV"` dominate `"cGCV"` for every problem —
  cGCV remains the default product selector.
- Not coordinate AIC/BIC λ search (lab prototype; IC formulas in
  `vignette("aic-bic")`).
- Not a substitute for choosing TT rank \(r\).

# See also

- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE for \(r\)
- `vignette("aic-bic", package = "TTPsplines")` — AIC / BIC from EDF
- `vignette("array-mode", package = "TTPsplines")` — cGCV / gGCV on product grids
- `vignette("getting-started", package = "TTPsplines")`
- `vignette("generalized", package = "TTPsplines")`

# References

Craven, P. and Wahba, G. (1979).
Smoothing noisy data with spline functions.
*Numerische Mathematik*, **31**, 377–403.

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
