---
title: "Generalized TTPsplines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized TTPsplines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Family-aware `auto`

With `optimizer = "auto"` (default):

| Family | Selected optimizer |
|--------|--------------------|
| Gaussian | `ALS` |
| Poisson | `PIRLS-ALS` |
| binomial | `LBFGS` |

```{r}
library(TTPsplines)
data(ishigami)
X <- as.matrix(ishigami[, c("x1", "x2", "x3")])
```

# Poisson

```{r poisson}
set.seed(2)
eta <- ishigami$f - mean(ishigami$f) + log(3)
y_p <- rpois(nrow(ishigami), exp(eta))
fit_p <- ttps(
  y_p, X, family = poisson(), rank = 2, k = 6, lambda = 1,
  control = tt_control(pirls_maxit = 20, backend = "R", compute_edf = FALSE)
)
c(used = fit_p$optimizer_used, reason = fit_p$optimizer_reason)
predict(fit_p, X[1:5, ], type = "response")
```

# Bernoulli

```{r bernoulli}
set.seed(3)
eta_b <- 1.2 * scale(ishigami$f)[, 1]
y_b <- rbinom(nrow(ishigami), 1, plogis(eta_b))
fit_b <- ttps(
  y_b, X, family = binomial(), rank = 2, k = 6, lambda = 5,
  control = tt_control(lbfgs_maxit = 150, backend = "R", compute_edf = FALSE)
)
c(used = fit_b$optimizer_used, reason = fit_b$optimizer_reason)
predict(fit_b, X[1:5, ], type = "response")
```

Automatic λ: `lambda = "cGCV"` (see `fit$lambda_boundary` if estimates hug
the search bounds), or `lambda = "CV"` for k-fold search under ALS / PIRLS
(`tt_control(cv_folds, cv_sweeps, cv_rule)`). Joint `"gGCV"` is Gaussian-only
— see `vignette("cgcv")`.

# Parametric and additive confounders (`linear=`, `smooth=`)

The joint TT surface on \(X\) can sit next to **unpenalized** covariates
(`linear=`) and **1D additive** smooths (`smooth=`), updated by
block-coordinate ALS / PIRLS. Typical use: day-of-week dummies, a time trend,
or a cyclic month effect — calendar structure that should **not** be folded
into the TT interaction surface.

Do **not** put an intercept column in `linear=` (`ttps` already estimates one).
`predict()` must receive matching `linear=` / `smooth=` newdata when those
terms were used at fit time. Unsupported for global LBFGS / GD / hybrids
(use ALS / PIRLS).

```{r linear-smooth}
set.seed(10)
n <- 150
X2 <- matrix(runif(n * 2), n, 2)
colnames(X2) <- c("x1", "x2")
z <- rnorm(n)
t <- seq(0, 1, length.out = n)
f_tt <- 0.4 * sin(2 * pi * X2[, 1]) * cos(2 * pi * X2[, 2])
f_sm <- sin(2 * pi * t)
y <- 1 + 1.5 * z + f_tt + f_sm + rnorm(n, 0, 0.2)

fit_ls <- ttps(
  y, X2,
  rank = 2, k = 6, lambda = 1,
  linear = cbind(z = z),
  smooth = list(
    time = list(x = t, bs = "ps", k = 12, m = 2, lambda = "cGCV")
  ),
  control = tt_control(max_sweeps = 8, compute_edf = FALSE, outer_maxit = 4)
)
c(beta_z = unname(fit_ls$beta), smooth_edf = fit_ls$smooth$time$edf)
head(predict(fit_ls, X2, linear = cbind(z = z), smooth = list(time = t)))
```

Cyclic margins use `bs = "cc"` with `period = c(xl, xr)` (e.g. hour of day).
Per-term `target_edf` is an alternative to `lambda` / `"cGCV"` on a smooth.

# Product grids: TT array vs dense GLAM

For a **full rectangular** response array there are two package paths:

| Path | Coefficients | When |
|------|--------------|------|
| `ttps(Y, axes = ..., array = TRUE)` | TT cores | Same TT model as scattered; faster Gram/RHS on the grid |
| `glam_fit_poisson()` | Dense \(\Theta\) | Classical Currie–Durbán–Eilers baseline (exposure tables) |

Walkthrough of `array = TRUE` (Gaussian and Poisson):
`vignette("array-mode")`. Dense GLAM Poisson:
`vignette("glam-poisson")` and `data(glam_poisson)`.

# See also

- `vignette("getting-started", package = "TTPsplines")`
- `vignette("array-mode", package = "TTPsplines")` — product grids (`array = TRUE`)
- `vignette("cgcv", package = "TTPsplines")` — cGCV / CV / gGCV for \(\lambda\)
- `vignette("rank-selection", package = "TTPsplines")`
- `vignette("glam-poisson", package = "TTPsplines")`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280. <https://doi.org/10.1111/j.1467-9868.2006.00543.x>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>

McCullagh, P. and Nelder, J. A. (1989).
*Generalized Linear Models* (2nd ed.). Chapman & Hall/CRC.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>
