---
title: "Generalized TTPsplines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized TTPsplines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Family-aware `auto`

With `optimizer = "auto"` (default):

| Family | Selected optimizer |
|--------|--------------------|
| Gaussian | `ALS` |
| Poisson | `PIRLS-ALS` |
| binomial | `LBFGS` |

```{r}
library(TTPsplines)
data(ishigami)
X <- as.matrix(ishigami[, c("x1", "x2", "x3")])
```

# Poisson

```{r poisson}
set.seed(2)
eta <- ishigami$f - mean(ishigami$f) + log(3)
y_p <- rpois(nrow(ishigami), exp(eta))
fit_p <- ttps(
  y_p, X, family = poisson(), rank = 2, k = 6, lambda = 1,
  control = tt_control(pirls_maxit = 20, backend = "R", compute_edf = FALSE)
)
c(used = fit_p$optimizer_used, reason = fit_p$optimizer_reason)
predict(fit_p, X[1:5, ], type = "response")
```

# Bernoulli

```{r bernoulli}
set.seed(3)
eta_b <- 1.2 * scale(ishigami$f)[, 1]
y_b <- rbinom(nrow(ishigami), 1, plogis(eta_b))
fit_b <- ttps(
  y_b, X, family = binomial(), rank = 2, k = 6, lambda = 5,
  control = tt_control(lbfgs_maxit = 150, backend = "R", compute_edf = FALSE)
)
c(used = fit_b$optimizer_used, reason = fit_b$optimizer_reason)
predict(fit_b, X[1:5, ], type = "response")
```

Automatic λ: `lambda = "cGCV"` (see `fit$lambda_boundary` if estimates hug
the search bounds). Future versions may add `"cFS"` / `"cREML"` via the same
argument.

# Grid Poisson via GLAM

For **regular arrays** (e.g. age × year counts with exposure), use
`glam_fit_poisson()` instead of scattering the grid into `ttps()` —
see `vignette("glam-poisson")` and `data(glam_poisson)`.

# See also

- `vignette("getting-started", package = "TTPsplines")`
- `vignette("cgcv", package = "TTPsplines")` — automatic \(\lambda\)
- `vignette("rank-selection", package = "TTPsplines")`
- `vignette("glam-poisson", package = "TTPsplines")`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280. <https://doi.org/10.1111/j.1467-9868.2006.00543.x>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>

McCullagh, P. and Nelder, J. A. (1989).
*Generalized Linear Models* (2nd ed.). Chapman & Hall/CRC.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>
