---
title: "Getting started with TTPsplines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with TTPsplines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Scattered data, tensor coefficients, TT compression

`TTPsplines` models a **single non-additive** smooth

\[
g(\mathbb{E}[Y\mid x]) = f(x_1,\ldots,x_d)
\]

with a tensor-product P-spline whose coefficient array \(\Theta\) is stored in
**Tensor-Train** format. Observations \(X\in\mathbb{R}^{n\times d}\) may be
arbitrarily scattered — **no data grid** is required.

Family-aware default solver (`optimizer = "auto"`): Gaussian → ALS,
Poisson → PIRLS-ALS, binomial → L-BFGS. Rank \(r\) is **not** chosen inside
`ttps()` (see `vignette("rank-selection")`); directional \(\lambda\) can
be fixed or selected by cGCV (`vignette("cgcv")`).

# Why these datasets?

Classic UQ response surfaces are convenient demos: known truth, strong
interactions, and increasing dimension. Packaged samples (each `n = 800`):

```{r}
library(TTPsplines)
data(ishigami)
data(sobol_g)
data(friedman)
```

| Dataset | \(d\) | Domain | Role |
|---------|------:|--------|------|
| `ishigami` | 3 | \([-\pi,\pi]^3\) | Strong interaction / anisotropy |
| `sobol_g` | 4 | \([0,1]^4\) | Declining coordinate importance |
| `friedman` | 5 | \([0,1]^5\) | Mixed interaction + main effects |

These are **regression surfaces**, not Sobol-index estimators. To redraw
samples with different `n` / noise, use `simulate_ishigami()`,
`simulate_sobol_g()`, and `simulate_friedman()`.

# Ishigami (\(d=3\))

The Ishigami surface admits an approximately low-rank functional form
(structural TT rank \(\lesssim 2\)), so `rank = 2` is a natural demo.
ALS at that minimal rank can still be **initialization-sensitive**: a bad
start may look like “\(r=2\) is inadequate” even when capacity is fine.
For scientific rank choice prefer `tt_rank_select(..., n_starts > 1)`
(`vignette("rank-selection")`); warm-start via `tt_truncate_rank()` is
documented there and in the Ishigami init audit under `inst/benchmarks/`.

```{r ishigami-fit}
data(ishigami)
X_i <- as.matrix(ishigami[, c("x1", "x2", "x3")])
fit_i <- ttps(
  ishigami$y, X_i,
  family = gaussian(),
  rank = 2, k = 8, lambda = 1,
  control = tt_control(max_sweeps = 10, backend = "R", compute_edf = TRUE,
                       seed = 1L)
)
summary(fit_i)
tt_complexity(fit_i)
c(
  requested = fit_i$optimizer_requested,
  used = fit_i$optimizer_used,
  reason = fit_i$optimizer_reason
)
```

True surface vs fitted predictor:

```{r ishigami-rmse}
rmse <- function(a, b) sqrt(mean((a - b)^2))
c(
  RMSE_y = rmse(fitted(fit_i), ishigami$y),
  RMSE_f = rmse(fitted(fit_i), ishigami$f)
)
```

Compression story on this fit:

```{r ishigami-layers}
c(
  full_tensor = fit_i$npar_dense,
  TT_stored = fit_i$npar_tt,
  TT_intrinsic = fit_i$npar_tt_intrinsic,
  EDF = fit_i$edf
)
```

# Sobol \(g\)-function (\(d=4\))

With \(a=(0,0.5,3,9)\), earlier coordinates dominate.

```{r sobol-fit}
data(sobol_g)
X_s <- as.matrix(sobol_g[, paste0("x", 1:4)])
fit_s <- ttps(
  sobol_g$y, X_s,
  family = gaussian(),
  rank = 2, k = 6, lambda = 1,
  control = tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE)
)
fit_s$optimizer_used
fit_s$compression_ratio
rmse(fitted(fit_s), sobol_g$f)
```

For anisotropic automatic \(\lambda\), see `vignette("cgcv")`
(`lambda = "cGCV"`, boundary diagnostics).

# Friedman #1 (\(d=5\))

```{r friedman-fit}
data(friedman)
X_f <- as.matrix(friedman[, paste0("x", 1:5)])
fit_f <- ttps(
  friedman$y, X_f,
  family = gaussian(),
  rank = 3, k = 5, lambda = 1,
  control = tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE)
)
summary(fit_f)
rmse(fitted(fit_f), friedman$f)
predict(fit_f, X_f[1:5, ], type = "response")
```

# Family-aware `auto` on a test surface

Same Ishigami covariates, binomial response from the latent surface:

```{r ishigami-bern}
set.seed(11)
eta <- 1.2 * scale(ishigami$f)[, 1]
y_b <- rbinom(nrow(ishigami), 1, plogis(eta))
fit_b <- ttps(
  y_b, X_i,
  family = binomial(),
  rank = 2, k = 6, lambda = 5,
  control = tt_control(lbfgs_maxit = 150, backend = "R", compute_edf = FALSE)
)
c(
  requested = fit_b$optimizer_requested,
  used = fit_b$optimizer_used,
  reason = fit_b$optimizer_reason
)
predict(fit_b, X_i[1:5, ], type = "response")
```

# Regenerating the packaged samples

```{r regen, eval = FALSE}
# Rscript data-raw/build_example_datasets.R
dat <- simulate_ishigami(n = 2000, sigma = 0.1, seed = 99)
fit <- ttps(dat$y, dat$X, rank = 3, k = 10, lambda = "cGCV")
```

# See also

- `vignette("generalized", package = "TTPsplines")` — Poisson / Bernoulli
- `vignette("cgcv", package = "TTPsplines")` — fixed vs conditional GCV for \(\lambda\)
- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE for \(r\)
- `vignette("aic-bic", package = "TTPsplines")` — AIC / BIC from linearized EDF
- `vignette("glam-vs-tt", package = "TTPsplines")` — full tensor vs TT
- `vignette("scalability", package = "TTPsplines")` — storage, timings, backends
- `vignette("glam-poisson", package = "TTPsplines")` — Currie–Durbán–Eilers Poisson
- `vignette("uncertainty", package = "TTPsplines")` — conditional SEs

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280. <https://doi.org/10.1111/j.1467-9868.2006.00543.x>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>

Friedman, J. H. (1991).
Multivariate adaptive regression splines.
*The Annals of Statistics*, **19**, 1–67.
<https://doi.org/10.1214/aos/1176347963>

Ishigami, T. and Homma, T. (1990).
An importance quantification technique in uncertainty analysis for computer
models. In *Proceedings of the ISUMA'90, First International Symposium on
Uncertainty Modeling and Analysis*, 398–403. IEEE.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>

Saltelli, A., Chan, K. and Scott, E. M. (eds) (2000).
*Sensitivity Analysis*. Wiley.
