---
title: "Getting started with TTPsplines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with TTPsplines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Why these datasets?

`TTPsplines` targets **non-additive** multidimensional smoothing on scattered
covariates. Classic UQ response surfaces are convenient demos: known truth,
strong interactions, and increasing dimension.

Packaged samples (each `n = 800`) load with the usual R idiom:

```{r}
library(TTPsplines)
data(ishigami)
data(sobol_g)
data(friedman)
```

| Dataset | \(d\) | Domain | Role |
|---------|------:|--------|------|
| `ishigami` | 3 | \([-\pi,\pi]^3\) | Strong interaction / anisotropy |
| `sobol_g` | 4 | \([0,1]^4\) | Declining coordinate importance |
| `friedman` | 5 | \([0,1]^5\) | Mixed interaction + main effects |

These are **regression surfaces**, not Sobol-index estimators. To redraw
samples with different `n` / noise, use `simulate_ishigami()`,
`simulate_sobol_g()`, and `simulate_friedman()`.

# Ishigami (\(d=3\))

```{r ishigami-fit}
data(ishigami)
X_i <- as.matrix(ishigami[, c("x1", "x2", "x3")])
fit_i <- ttpspline(
  ishigami$y, X_i,
  family = gaussian(),
  rank = 2, k = 8, lambda = 1,
  control = tt_control(max_sweeps = 10, backend = "R", compute_edf = TRUE)
)
summary(fit_i)
tt_complexity(fit_i)
```

True surface vs fitted predictor:

```{r ishigami-rmse}
rmse <- function(a, b) sqrt(mean((a - b)^2))
c(
  RMSE_y = rmse(fitted(fit_i), ishigami$y),
  RMSE_f = rmse(fitted(fit_i), ishigami$f)
)
```

Compression story on this fit:

```{r ishigami-layers}
c(
  full_tensor = fit_i$npar_dense,
  TT_stored = fit_i$npar_tt,
  TT_intrinsic = fit_i$npar_tt_intrinsic,
  EDF = fit_i$edf
)
```

# Sobol \(g\)-function (\(d=4\))

With \(a=(0,0.5,3,9)\), earlier coordinates dominate.

```{r sobol-fit}
data(sobol_g)
X_s <- as.matrix(sobol_g[, paste0("x", 1:4)])
fit_s <- ttpspline(
  sobol_g$y, X_s,
  family = gaussian(),
  rank = 2, k = 6, lambda = 1,
  control = tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE)
)
fit_s$optimizer_used
fit_s$compression_ratio
rmse(fitted(fit_s), sobol_g$f)
```

Anisotropic cGCV often pushes directional \(\lambda\) apart (and sometimes onto
search bounds — see `fit$lambda_boundary` when `lambda = "cGCV"`).

```{r sobol-cgcv, warning=TRUE}
fit_s_cgcv <- ttpspline(
  sobol_g$y, X_s,
  family = gaussian(),
  rank = 2, k = 6, lambda = "cGCV",
  control = tt_control(
    max_sweeps = 8, backend = "R", compute_edf = FALSE,
    warn_lambda_boundary = TRUE
  )
)
fit_s_cgcv$lambda
fit_s_cgcv$lambda_boundary
```

# Friedman #1 (\(d=5\))

```{r friedman-fit}
data(friedman)
X_f <- as.matrix(friedman[, paste0("x", 1:5)])
fit_f <- ttpspline(
  friedman$y, X_f,
  family = gaussian(),
  rank = 3, k = 5, lambda = 1,
  control = tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE)
)
summary(fit_f)
rmse(fitted(fit_f), friedman$f)
predict(fit_f, X_f[1:5, ], type = "response")
```

# Family-aware `auto` on a test surface

Same Ishigami covariates, binomial response from the latent surface:

```{r ishigami-bern}
set.seed(11)
eta <- 1.2 * scale(ishigami$f)[, 1]
y_b <- rbinom(nrow(ishigami), 1, plogis(eta))
fit_b <- ttpspline(
  y_b, X_i,
  family = binomial(),
  rank = 2, k = 6, lambda = 5,
  control = tt_control(lbfgs_maxit = 150, backend = "R", compute_edf = FALSE)
)
c(
  requested = fit_b$optimizer_requested,
  used = fit_b$optimizer_used,
  reason = fit_b$optimizer_reason
)
predict(fit_b, X_i[1:5, ], type = "response")
```

# Regenerating the packaged samples

```{r regen, eval = FALSE}
# Rscript data-raw/build_example_datasets.R
dat <- simulate_ishigami(n = 2000, sigma = 0.1, seed = 99)
fit <- ttpspline(dat$y, dat$X, rank = 3, k = 10, lambda = "cGCV")
```
