---
title: "GLAM Poisson (Currie array methods)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLAM Poisson (Currie array methods)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Grid arrays vs scattered TT

`ttpspline()` targets **scattered** covariates with a Tensor-Train coefficient
array. When responses live on a **regular multiway grid**, Currie–Durbán–Eilers
**GLAM** (generalized linear array models) evaluate the same tensor-product
P-spline without forming the Kronecker design, via rotated-\(H\) algebra.

This vignette shows fixed-\(\lambda\) **Poisson** GLAM with an exposure offset
— the classical age × year mortality-style setting.

```{r}
library(TTPsplines)
data(glam_poisson)
str(glam_poisson[c("Y", "exposure", "age", "year")], vec.len = 2)
```

# Fit

```{r fit}
bb <- glam_grid_bases(
  list(age = glam_poisson$age, year = glam_poisson$year),
  k = 10
)
fit <- glam_fit_poisson(
  glam_poisson$Y, bb$B,
  lambda = c(10, 1),
  offset = log(glam_poisson$exposure)
)
c(
  npar = fit$npar,
  n_pirls = fit$n_pirls,
  deviance = round(fit$deviance, 2)
)
```

Mean counts vs truth (log scale):

```{r rmse}
sqrt(mean((log(as.numeric(fit$mu)) - log(as.numeric(glam_poisson$mu)))^2))
```

Redraw a smaller grid on the fly with `simulate_glam_poisson()`.
For scattered Poisson / Bernoulli, see the *Generalized TTPsplines* vignette.
