---
title: "GLAM Poisson (Currie–Durbán–Eilers)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLAM Poisson (Currie–Durbán–Eilers)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Grid arrays vs scattered TT

Classical GLAM (Currie, Durbán and Eilers, 2006) exploits **array structure
in the observations and the design**: responses live on a product grid
(e.g.\ age × year), and Kronecker / rotated-\(H\) algebra avoids forming the
full tensor-product design. The coefficient array \(\Theta\) is still dense
of size \(\prod_j k_j\).

`ttps()` / TTPsplines exploits **tensor structure in the coefficient
representation** (TT cores) and allows **arbitrarily scattered** covariates —
not “GLAM with Tensor-Train cores,” but a complementary use of tensor
structure.

This vignette shows fixed-\(\lambda\) **Poisson** GLAM with an exposure offset
— the classical age × year mortality-style setting of Currie, Durbán and
Eilers (2006).

```{r}
library(TTPsplines)
data(glam_poisson)
str(glam_poisson[c("Y", "exposure", "age", "year")], vec.len = 2)
```

# Fit

```{r fit}
bb <- glam_grid_bases(
  list(age = glam_poisson$age, year = glam_poisson$year),
  k = 10
)
fit <- glam_fit_poisson(
  glam_poisson$Y, bb$B,
  lambda = c(10, 1),
  offset = log(glam_poisson$exposure)
)
c(
  npar = fit$npar,
  n_pirls = fit$n_pirls,
  deviance = round(fit$deviance, 2)
)
```

Mean counts vs truth (log scale):

```{r rmse}
rmse_log <- function(a, b) sqrt(mean((log(as.numeric(a)) - log(as.numeric(b)))^2))
rmse_log(fit$mu, glam_poisson$mu)
```

# Comparison with TTPsplines

The same age × year cells can be treated as **scattered** observations and fit
with `ttps()` (Poisson / `PIRLS-ALS`). Currie–Durbán–Eilers GLAM keeps the
array algebra; TT compresses the coefficient tensor. Both accept an exposure
`offset` on the linear predictor
(\(\eta = \mathrm{offset} + \alpha + f(x)\)).

```{r tt-vs-glam}
X <- as.matrix(expand.grid(
  age = glam_poisson$age,
  year = glam_poisson$year
))
y <- as.numeric(glam_poisson$Y)
mu_true <- as.numeric(glam_poisson$mu)
off <- log(as.numeric(glam_poisson$exposure))

fit_tt <- ttps(
  y, X,
  family = poisson(),
  rank = 3, k = 10, lambda = c(10, 1),
  offset = off,
  control = tt_control(
    pirls_maxit = 20, max_sweeps = 8,
    backend = "R", compute_edf = FALSE
  )
)

cmp <- data.frame(
  method = c("GLAM (Currie–Durbán–Eilers)", "TTPsplines (rank 3)"),
  npar = c(fit$npar, fit_tt$npar_tt),
  npar_dense = c(fit$npar, fit_tt$npar_dense),
  compression = c(1, round(fit_tt$compression_ratio, 2)),
  RMSE_log_mu = c(
    rmse_log(fit$mu, mu_true),
    rmse_log(fitted(fit_tt), mu_true)
  ),
  RMSE_vs_GLAM = c(
    0,
    rmse_log(fitted(fit_tt), fit$mu)
  ),
  optimizer = c(fit$method, fit_tt$optimizer_used),
  stringsAsFactors = FALSE
)
print(cmp, row.names = FALSE)
```

```{r tt-complexity}
tt_complexity(fit_tt)
```

GLAM uses the full \(k\times k\) coefficient array; TT stores fewer parameters
at modest rank while staying close to the GLAM mean surface on this example.
For scattered Poisson / Bernoulli (no grid), prefer `ttps()` directly —
see `vignette("generalized")`. For Gaussian GLAM vs TT with CV ranks and
separated timings, see `vignette("glam-vs-tt")`.

Redraw a smaller grid on the fly with `simulate_glam_poisson()`.

# See also

- `vignette("generalized", package = "TTPsplines")` — scattered Poisson / Bernoulli
- `vignette("glam-vs-tt", package = "TTPsplines")` — full tensor vs TT
- `vignette("cgcv", package = "TTPsplines")` — automatic \(\lambda\)
- `vignette("rank-selection", package = "TTPsplines")`
- `vignette("getting-started", package = "TTPsplines")`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280. <https://doi.org/10.1111/j.1467-9868.2006.00543.x>

Eilers, P. H. C., Currie, I. D. and Durban, M. (2006).
Fast and compact smoothing on large multidimensional grids.
*Computational Statistics & Data Analysis*, **50**, 61–76.
<https://doi.org/10.1016/j.csda.2004.07.008>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>
