---
title: "Gaussian GLAM vs TTPsplines (d = 3, 5, 7)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian GLAM vs TTPsplines (d = 3, 5, 7)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Setup

Grid-based Currie–Durbán–Eilers **GLAM** vs **TT** with data-driven ranks
(`TT-minCV`, `TT-1SE`) and a simulation-only `TT-oracle`.

**Timings are separated:** `fit_time` (final model) vs
`rank_selection_time` (CV) vs `total_procedure_time` (sum). Oracle probe
cost is **not** folded into practical totals.

The full narrative lives in `vignette("glam-vs-tt")`.

```{r}
library(TTPsplines)
cmp <- compare_glam_tt_gaussian(
  d = 3, n_grid = 10, k = 6, ranks = 1:3,
  lambda = 1, folds = 3, max_sweeps = 4, seed = 1, n_test = 500
)
print(cmp)
summarize_glam_tt_compare(cmp)
```

See also `vignette("glam-vs-tt")`, `vignette("rank-selection")`,
`vignette("glam-poisson")`.

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006). *JRSS-B*.

Oseledets, I. V. (2011). *SIAM J. Sci. Comput.*
