---
title: "Full tensor vs Tensor-Train smoothing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Full tensor vs Tensor-Train smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Motivation

A tensor-product P-spline stores about \(k^d\) coefficients. The TT
representation stores on the order of \(O(d k r^2)\) parameters for uniform
interior ranks — a scaling statement, not an exact identity.

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

Rank is **structural capacity**; \(\lambda\) is **smoothness**. Rank is
**not** chosen inside `ttpspline()`:

```r
sel <- tt_rank_select(...)
fit <- tt_rank_refit(sel)
```

# Timing: be honest with the referee

Selecting \(r\) by \(K\)-fold CV costs many fits. The comparison therefore
reports **separately**:

| Column | Meaning |
|--------|---------|
| `fit_time` | Final model fit (rank known) |
| `rank_selection_time` | Cost of `tt_rank_select()` |
| `total_procedure_time` | `fit_time + rank_selection_time` |
| `oracle_probe_time` | Simulation-only cost of probing all ranks (**not** in totals) |

```{r}
library(TTPsplines)
```

# Low-\(d\) reference (\(d=3\))

GLAM is cheap; TT may add selection overhead but compresses.

```{r d3}
cmp3 <- compare_glam_tt_gaussian(
  d = 3, n_grid = c(10, 10, 10), k = 6, ranks = 1:4,
  lambda = 1, folds = 4, rank_rule = "both",
  max_sweeps = 5, seed = 1, detail_ranks = TRUE, n_test = 800
)
print(cmp3)
```

```{r d3-plots, fig.width = 7, fig.height = 3.5}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(cmp3, type = "rmse_rank")
plot(cmp3, type = "tradeoff")
par(op)
```

# Higher-\(d\) case (\(d=5\))

```{r d5}
cmp5 <- compare_glam_tt_gaussian(
  d = 5, n_grid = 6, k = 5, ranks = 1:4,
  lambda = 1, folds = 4, max_sweeps = 5, seed = 1,
  detail_ranks = TRUE, n_test = 1000
)
analyze_glam_tt_rank_select(cmp5)
```

# Oracle vs data-driven ranks

| Method | Role |
|--------|------|
| `GLAM` | Full tensor reference |
| `TT-oracle` | \(\arg\min_r\mathrm{RMSE}_{\mathrm{test}}(\hat f_r,f_{\mathrm{true}})\) — **simulation only** |
| `TT-minCV` | Minimum CV on training folds |
| `TT-1SE` | Parsimonious CV rule (recommended practical default) |

Never choose rank using truth on real data.

# Practical recommendation

```r
sel <- tt_rank_select(y, X, ranks = 1:5, lambda = 1, folds = 5, rule = "1se")
fit <- tt_rank_refit(sel)
```

See also `vignette("rank-selection")` for the selection API alone,
`vignette("glam-gaussian")` for a shorter GLAM-vs-TT demo, and
`vignette("rank-compression")` for complexity layers.

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006). *JRSS-B*.

Oseledets, I. V. (2011). *SIAM J. Sci. Comput.*
