---
title: "Gaussian GLAM vs TT (quick start)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian GLAM vs TT (quick start)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Setup

Short demo of grid **GLAM** vs **TT** with data-driven ranks
(`TT-minCV`, `TT-1SE`) and a simulation-only `TT-oracle`.

**Timings are separated:** `fit_time` | `rank_selection_time` |
`total_procedure_time`. Oracle probe cost is **not** in practical totals.

Full narrative (plots, \(d=5\), accuracy–compression): `vignette("glam-vs-tt")`.

```{r}
library(TTPsplines)
cmp <- compare_glam_tt_gaussian(
  d = 3, n_grid = 10, k = 6, ranks = 1:3,
  lambda = 1, folds = 3, max_sweeps = 4, seed = 1, n_test = 500
)
print(cmp)
summarize_glam_tt_compare(cmp)
```

# See also

- `vignette("glam-vs-tt", package = "TTPsplines")`
- `vignette("rank-selection", package = "TTPsplines")`
- `vignette("glam-poisson", package = "TTPsplines")`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006). *JRSS-B*.

Oseledets, I. V. (2011). *SIAM J. Sci. Comput.*
