---
title: "Gaussian GLAM vs TTPsplines (d = 3 and 5)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gaussian GLAM vs TTPsplines (d = 3 and 5)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Setup

On a **regular grid**, Currie–Durbán–Eilers **GLAM** fits a full tensor-product
P-spline coefficient array with rotated-\(H\) algebra. `ttpspline()` fits the
same kind of surface after vectorizing the grid, but stores the coefficients in
**Tensor-Train** format.

This vignette uses fixed \(\lambda\) (no cGCV) so the comparison is about
**compression, accuracy and time**, not smoothing-parameter selection.
Helper: `compare_glam_tt_gaussian()`.

```{r}
library(TTPsplines)
cols <- c("method", "npar", "npar_dense", "compression",
          "rmse_truth", "rmse_y", "rmse_vs_glam", "time_s", "time_vs_glam")
```

# \(d = 3\)

Default grid \(14\times 12\times 10\), \(k=6\) (dense \(k^d = 216\)).

```{r d3}
tab3 <- compare_glam_tt_gaussian(d = 3, ranks = 1:4, max_sweeps = 10)
print(tab3[, cols], digits = 3, row.names = FALSE)
```

```{r d3-plot, fig.width = 7, fig.height = 3.5}
tt3 <- tab3[!is.na(tab3$rank), ]
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(tt3$rank, tt3$rmse_truth, type = "b", pch = 19,
     xlab = "TT rank", ylab = "RMSE vs truth",
     main = "d = 3: accuracy",
     ylim = range(c(tt3$rmse_truth, tab3$rmse_truth[1])))
abline(h = tab3$rmse_truth[1], col = 2, lty = 2)
legend("topright", c("TT", "GLAM"), col = c(1, 2), lty = c(1, 2),
       pch = c(19, NA), bty = "n", cex = 0.8)
plot(tt3$rank, tt3$npar, type = "b", pch = 19,
     xlab = "TT rank", ylab = "Stored parameters",
     main = "d = 3: compression",
     ylim = range(c(tt3$npar, tab3$npar[1])))
abline(h = tab3$npar[1], col = 2, lty = 2)
par(op)
```

# \(d = 5\)

Default grid \(6^5 = 7776\) cells, \(k=5\) (dense \(k^d = 3125\)). Full GLAM
still builds a dense \(3125\times 3125\) system; TT stays linear in \(d\) at
fixed rank.

```{r d5}
tab5 <- compare_glam_tt_gaussian(d = 5, ranks = 1:3, max_sweeps = 8)
print(tab5[, cols], digits = 3, row.names = FALSE)
```

```{r d5-plot, fig.width = 7, fig.height = 3.5}
tt5 <- tab5[!is.na(tab5$rank), ]
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(tt5$rank, tt5$rmse_truth, type = "b", pch = 19,
     xlab = "TT rank", ylab = "RMSE vs truth",
     main = "d = 5: accuracy",
     ylim = range(c(tt5$rmse_truth, tab5$rmse_truth[1])))
abline(h = tab5$rmse_truth[1], col = 2, lty = 2)
legend("topright", c("TT", "GLAM"), col = c(1, 2), lty = c(1, 2),
       pch = c(19, NA), bty = "n", cex = 0.8)
plot(tt5$rank, tt5$time_s, type = "b", pch = 19,
     xlab = "TT rank", ylab = "Time (s)",
     main = "d = 5: wall time",
     ylim = range(c(tt5$time_s, tab5$time_s[1])))
abline(h = tab5$time_s[1], col = 2, lty = 2)
par(op)
```

# How to read the table

| Column | Meaning |
|--------|---------|
| `npar` | Stored parameters (GLAM = full \(k^d\); TT = TT cores) |
| `compression` | `npar_dense / npar` |
| `rmse_truth` | RMSE vs the known smooth |
| `rmse_y` | RMSE vs noisy data |
| `rmse_vs_glam` | RMSE of TT means vs GLAM means |
| `time_s` | Wall time |
| `time_vs_glam` | Time relative to GLAM on the same grid |

Typical pattern: at small \(d\), GLAM is cheap and a useful dense baseline; as
\(d\) grows, TT wins on **storage and time**, and a moderate rank often matches
or beats GLAM accuracy under the same fixed \(\lambda\).

For Poisson age × year with exposure, see `vignette("glam-poisson")`.

# See also

- `vignette("glam-poisson", package = "TTPsplines")`
- `vignette("rank-compression", package = "TTPsplines")`
- `vignette("getting-started", package = "TTPsplines")`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280. <https://doi.org/10.1111/j.1467-9868.2006.00543.x>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>
