---
title: "Introduction to TTPsplines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to TTPsplines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Scattered data, tensor coefficients, TT compression

`TTPsplines` models a **single non-additive** smooth

\[
g(\mathbb{E}[Y\mid x]) = f(x_1,\ldots,x_d)
\]

with a tensor-product P-spline whose coefficient array \(\Theta\) is stored in
**Tensor-Train** format. Observations \(X\in\mathbb{R}^{n\times d}\) may be
arbitrarily scattered — **no data grid** is required.

```{r}
library(TTPsplines)
data(ishigami)
X <- as.matrix(ishigami[, c("x1", "x2", "x3")])
fit <- ttpspline(
  ishigami$y, X,
  family = gaussian(),
  rank = 2, k = 8, lambda = 1,
  control = tt_control(max_sweeps = 10, backend = "R", compute_edf = TRUE)
)
summary(fit)
tt_complexity(fit)
```

Family-aware default solver (`optimizer = "auto"`):

```{r}
c(
  requested = fit$optimizer_requested,
  used = fit$optimizer_used,
  reason = fit$optimizer_reason
)
```

Also available: `data(sobol_g)`, `data(friedman)`. See *Getting started with
TTPsplines* for Ishigami / Sobol-\(g\) / Friedman demos, Poisson / Bernoulli
usage in *Generalized TTPsplines*, and Currie-style grid Poisson in
*GLAM Poisson (Currie array methods)* (`data(glam_poisson)`).

# See also

- `vignette("getting-started", package = "TTPsplines")`
- `vignette("generalized", package = "TTPsplines")`
- `vignette("glam-poisson", package = "TTPsplines")`
- `vignette("rank-compression", package = "TTPsplines")`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280. <https://doi.org/10.1111/j.1467-9868.2006.00543.x>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>
