---
title: "Margin Activity Path"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Margin Activity Path}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3.2
)
```

# Why screen margins?

`ttps()` fits a **joint** non-additive smooth on all columns of `X`.
When many margins are null, conditional GCV on the full \(d\) can still
assign small \(\lambda_j\) to noise directions and hurt holdout risk.
**Margin Activity Path** ranks covariates by how much the fitted surface
moves when each margin varies (partial range) along an isotropic
\(\lambda\) path, then chooses a nested top-\(m\) subset by CV
(default: one-SE rule). Selected margins are refit with `lambda = "cGCV"`.

This is **not** maximum a posteriori. Prefer the full name
*Margin Activity Path* over the ambiguous acronym "MAP".

```{r}
library(TTPsplines)
```

Formal protocol (lab): `inst/benchmarks/margin_path/PROTOCOL_MAP.md`.

# Toy example (\(d=6\), two active sins)

```{r data}
set.seed(1)
n <- 200
d <- 6
X <- matrix(runif(n * d), n, d)
colnames(X) <- paste0("x", seq_len(d))
y <- sin(2 * pi * X[, 1]) + sin(2 * pi * X[, 2]) +
  stats::rnorm(n, sd = 0.3)
```

```{r path}
ctrl <- tt_control(
  max_sweeps = 5, compute_edf = FALSE, outer_maxit = 5, backend = "R"
)
path <- tt_margin_activity_path(
  y, X,
  rank = 2, k = 5,
  lambda_path = 10^c(1.5, 1, 0.5, 0, -0.5),
  select = "1se",
  folds = 3,
  seed = 1,
  control = ctrl,
  verbose = FALSE
)
path
path$selected_names
```

```{r plot, fig.width = 8, fig.height = 3}
plot(path)
```

# Compare to full-\(d\) cGCV

```{r holdout}
set.seed(2)
Xte <- matrix(runif(n * d), n, d)
colnames(Xte) <- colnames(X)
yte <- sin(2 * pi * Xte[, 1]) + sin(2 * pi * Xte[, 2]) +
  stats::rnorm(n, sd = 0.3)

fit_full <- ttps(
  y, X, rank = 2, k = 5, lambda = "cGCV", control = ctrl
)
mse <- function(fit, Xnew, ynew) {
  mean((as.numeric(predict(fit, Xnew)) - ynew)^2)
}

data.frame(
  method = c("full TT + cGCV", "Margin Activity Path + cGCV"),
  n_margins = c(d, length(path$selected)),
  test_MSE = c(
    mse(fit_full, Xte, yte),
    mse(path$fit, Xte[, path$selected, drop = FALSE], yte)
  )
)
```

# Scores only (no CV)

```{r scores-only}
path0 <- tt_margin_activity_path(
  y, X, rank = 2, k = 5,
  lambda_path = 10^c(1, 0, -1),
  select = "none",
  control = ctrl,
  verbose = FALSE
)
round(sort(path0$scores, decreasing = TRUE), 3)
```

# Relation to other helpers

| Helper | Role |
|--------|------|
| `ttps(..., lambda = "cGCV")` | Smooth on a **given** column set |
| `tt_rank_select()` | Choose TT **rank** \(r\) by CV |
| `tt_margin_activity_path()` | Choose **which margins** to keep |
| `tt_margin_drop_test()` | Leave-one-out / permute test to **drop** margins |
| `tt_edf(fit)` | Joint EDF + block `edf_margin` (sums) + `edf_margin_cond` (cGCV diag.) |

Keep \(r \neq \lambda \neq\) margin set conceptually distinct. After screening,
you may still run `tt_rank_select()` on the selected columns.
`fit$edf_margin` partitions joint `fit$edf` via \(\operatorname{tr}(H_{kk})\);
`fit$edf_margin_cond` is the ALS/cGCV conditional trace (not additive).

# Optional: drop test

```{r drop-test}
ctrl_edf <- tt_control(
  max_sweeps = 5, compute_edf = TRUE, outer_maxit = 5, backend = "R",
  seed = 1
)
tst <- tt_margin_drop_test(
  y, X, rank = 2, k = 5, lambda = 1, method = "nested",
  alpha = 0.05, seed = 1, control = ctrl_edf, verbose = FALSE
)
tst
tst$drop_candidate_names
```

`method = "permute"` is slower but closer to a proper null for each margin.

# Fully reproducible script

From the package source tree:

```r
Rscript inst/examples/reprex_margin_activity_path.R
```

Fixed seeds `20260817` / `20260818`, `backend = "R"`, and CSV outputs under
`inst/examples/margin_activity_path_figs/reprex_*.csv`.
