---
title: "Choosing the Tensor-Train rank"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the Tensor-Train rank}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Setup

`ttps()` **never** chooses the TT rank for you: `rank = 2` means
exactly rank 2. Predictive selection is a separate step:

```{r}
library(TTPsplines)
data(friedman)
# short subsample keeps the vignette install-time friendly
idx <- seq_len(250)
X <- as.matrix(friedman[idx, paste0("x", 1:5)])
y <- friedman$y[idx]
```

# Complexity layers

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

| Layer | Meaning |
|-------|---------|
| \(r\) | **structural capacity** (TT rank) |
| \(\lambda\) | **directional smoothness** (fixed or cGCV) |
| EDF | **effective flexibility** after P-spline penalization |
| \(k^d\) | Full tensor-product coefficients |
| TT stored | Packed core entries (\(\approx O(dkr^2)\) uniform ranks) |
| TT intrinsic | Stored minus gauge redundancy |

Increasing \(r\) enlarges the TT manifold (more stored parameters, less
compression). Smoothing \(\lambda\) is a different knob
(`vignette("cgcv")`).

```{r complexity}
data(ishigami)
X_i <- as.matrix(ishigami[, c("x1", "x2", "x3")])
fit0 <- ttps(
  ishigami$y, X_i, rank = 2, k = 8, lambda = 1,
  control = tt_control(max_sweeps = 6, compute_edf = TRUE)
)
tt_complexity(fit0)
```

# Cross-validation + 1-SE

`tt_rank_select()` uses a **single** K-fold split shared across candidate
ranks, a family-aware validation loss (RMSE for Gaussian), then:

- **minimum-CV** rank \(r_{\min}\);
- **1-SE** rank: the *smallest* \(r\) whose mean CV error is within one
  fold-SE of the minimum.

Optional **`n_starts`**: several random TT initializations per fold×rank;
the fold keeps the start with best training objective (preferring
converged fits). Use this when low-rank ALS is init-sensitive — Ishigami at
\(r=2\) can look “bad” with one start even though capacity is adequate.
`summary(sel)` then reports `objective_best`, `objective_median`,
`start_gap` (= median − best), and `start_convergence_rate`.

The 1-SE rule selects the smallest candidate rank whose cross-validated
error lies within one standard error of the minimum. This is a
**parsimonious predictive choice**, not a formal test of a “true” tensor
rank.

```{r select}
sel <- tt_rank_select(
  y, X,
  ranks = 1:4,
  family = gaussian(),
  k = 5,
  lambda = 1,          # fixed λ for a fast vignette build
  folds = 4,
  n_starts = 1L,       # raise for init-sensitive problems
  rule = "1se",
  seed = 1,
  control = tt_control(max_sweeps = 5, compute_edf = FALSE)
)
sel
```

```{r plot, fig.width = 6, fig.height = 4}
plot(sel)                 # CV error ± SE
plot(sel, type = "tradeoff")
```

```{r refit}
fit <- tt_rank_refit(sel)
summary(fit)
```

# Optional: cGCV inside folds

When `lambda = "cGCV"`, each fold selects \(\lambda\) **only on its
training rows**, then scores the held-out fold. Validation data never
enter cGCV. Details: `vignette("cgcv")`.

```{r cgcv, eval = FALSE}
sel_cgcv <- tt_rank_select(
  y, X, ranks = 1:4, k = 5, lambda = "cGCV",
  folds = 4, rule = "1se", seed = 1,
  control = tt_control(max_sweeps = 5, compute_edf = FALSE)
)
```

# In-sample rank sweep (diagnostic)

`tt_rank_profile()` refits on the **full** sample at each rank and reports
deviance / `npar`. This is **not** cross-validation.

```{r profile, fig.width = 6, fig.height = 3.5}
prof <- tt_rank_profile(
  ishigami$y, X_i, ranks = 1:4, k = 6, lambda = 1,
  control = tt_control(max_sweeps = 5, compute_edf = FALSE)
)
print(prof)
plot(prof)
```

# What this is not

- Not a likelihood-ratio / bootstrap test of TT rank.
- Not adaptive TT-SVD / greedy rank expansion.
- Not oracle \(r\) from simulation truth (`RMSE` vs \(f_{\mathrm{true}}\)).
- Not AIC/BIC rank selection (`vignette("aic-bic")`).

Oracle ranks are useful **only** in simulation to audit CV; they are not
part of the user API.

# See also

- `vignette("cgcv", package = "TTPsplines")` — choosing \(\lambda\)
- `vignette("aic-bic", package = "TTPsplines")` — in-sample AIC / BIC (EDF)
- `vignette("scalability", package = "TTPsplines")` — storage and timings
- `vignette("glam-vs-tt", package = "TTPsplines")` — full tensor vs TT
- `vignette("getting-started", package = "TTPsplines")`

# References

Hastie, T., Tibshirani, R. and Friedman, J. (2009).
*The Elements of Statistical Learning* (1-SE rule context).
Springer.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
