---
title: "Rank and compression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rank and compression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Tensor-Train (TT) ranks control how many parameters are stored relative to the
full \(p^d\) coefficient tensor. `tt_complexity(fit)` reports dense vs TT
counts and EDF layers; `tt_rank_profile()` sweeps ranks at fixed \(\lambda\).

```{r}
# prof <- tt_rank_profile(y, X, ranks = c(1, 2, 3, 4), lambda = 1)
# plot(prof)
# tt_complexity(fit)
```

# See also

- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE rank choice
- `vignette("getting-started", package = "TTPsplines")`
- `vignette("introduction", package = "TTPsplines")`

# References

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>
