---
title: "Choosing the Tensor-Train rank"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the Tensor-Train rank}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Setup

`ttpspline()` **never** chooses the TT rank for you: `rank = 2` means
exactly rank 2. Predictive selection is a separate step:

```{r}
library(TTPsplines)
data(friedman)
# short subsample keeps the vignette install-time friendly
idx <- seq_len(250)
X <- as.matrix(friedman[idx, paste0("x", 1:5)])
y <- friedman$y[idx]
```

# Three complexity layers

| Symbol | Role |
|--------|------|
| \(r\) | **structural capacity** (TT rank) |
| \(\lambda\) | **directional smoothness** |
| EDF | **effective fitted flexibility** after penalization |

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

Increasing \(r\) enlarges the TT manifold (more stored parameters, less
compression). It can improve prediction, but gains are often marginal.
Smoothing \(\lambda\) (fixed or cGCV) is a different knob.

# Cross-validation + 1-SE

`tt_rank_select()` uses a **single** K-fold split shared across candidate
ranks, a family-aware validation loss (RMSE for Gaussian), then:

- **minimum-CV** rank \(r_{\min}\);
- **1-SE** rank: the *smallest* \(r\) whose mean CV error is within one
  fold-SE of the minimum.

The 1-SE rule selects the smallest candidate rank whose cross-validated
error lies within one standard error of the minimum. This is a
**parsimonious predictive choice**, not a formal test of a “true” tensor
rank.

```{r select}
sel <- tt_rank_select(
  y, X,
  ranks = 1:4,
  family = gaussian(),
  k = 5,
  lambda = 1,          # fixed λ for a fast vignette build
  folds = 4,
  rule = "1se",
  seed = 1,
  control = tt_control(max_sweeps = 5, compute_edf = FALSE)
)
sel
```

```{r plot, fig.width = 6, fig.height = 4}
plot(sel)                 # CV error ± SE
plot(sel, type = "tradeoff")
```

```{r refit}
fit <- tt_rank_refit(sel)
summary(fit)
```

# Optional: cGCV inside folds

When `lambda = "cGCV"`, each fold selects \(\lambda\) **only on its
training rows**, then scores the held-out fold. Validation data never
enter cGCV.

```{r cgcv, eval = FALSE}
sel_cgcv <- tt_rank_select(
  y, X, ranks = 1:4, k = 5, lambda = "cGCV",
  folds = 4, rule = "1se", seed = 1,
  control = tt_control(max_sweeps = 5, compute_edf = FALSE)
)
```

# What this is not

- Not a likelihood-ratio / bootstrap test of TT rank.
- Not adaptive TT-SVD / greedy rank expansion.
- Not oracle \(r\) from simulation truth (`RMSE` vs \(f_{\mathrm{true}}\)).

Oracle ranks are useful **only** in simulation to audit CV; they are not
part of the user API.

# See also

- `vignette("glam-vs-tt", package = "TTPsplines")`
- `vignette("rank-compression", package = "TTPsplines")`
- `vignette("getting-started", package = "TTPsplines")`
- `tt_rank_profile()` — in-sample rank sweep (not CV)

# References

Hastie, T., Tibshirani, R. and Friedman, J. (2009).
*The Elements of Statistical Learning* (1-SE rule context).
Springer.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
