---
title: "Rank and compression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rank and compression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Complexity layers

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

| Layer | Meaning |
|-------|---------|
| \(k^d\) | Full tensor-product coefficients |
| TT stored | Entries in the cores (\(\approx O(dkr^2)\) for uniform interior ranks) |
| TT intrinsic | Stored minus gauge redundancy |
| EDF | Effective flexibility **after** P-spline penalization |

`ttpspline(..., rank = r)` uses that **exact** rank. For predictive choice of
\(r\), see `vignette("rank-selection")`.

```{r}
library(TTPsplines)
data(ishigami)
X <- as.matrix(ishigami[, c("x1", "x2", "x3")])
fit <- ttpspline(
  ishigami$y, X, rank = 2, k = 8, lambda = 1,
  control = tt_control(max_sweeps = 8, compute_edf = TRUE)
)
tt_complexity(fit)
```

# In-sample rank sweep (diagnostic)

`tt_rank_profile()` refits on the **full** sample at each rank and reports
deviance / `npar`. This is **not** cross-validation.

```{r profile, fig.width = 6, fig.height = 3.5}
prof <- tt_rank_profile(
  ishigami$y, X, ranks = 1:4, k = 6, lambda = 1,
  control = tt_control(max_sweeps = 6, compute_edf = FALSE)
)
print(prof)
plot(prof)
```

# See also

- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE
- `vignette("glam-vs-tt", package = "TTPsplines")` — full tensor vs TT
- `vignette("getting-started", package = "TTPsplines")`

# References

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
