---
title: "Scalability of TT P-splines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scalability of TT P-splines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# What this vignette is

Tensor-product P-splines store about \(k^d\) coefficients. A Tensor-Train
(TT) factorization of the **coefficient** array stores on the order of
\(O(d k r^2)\) parameters for uniform interior ranks — a **scaling**
statement, not an exact identity.

This note is about **storage and wall time**, not predictive rank choice
(`vignette("rank-selection")`) or uncertainty
(`vignette("uncertainty")`).

```{r}
library(TTPsplines)
```

# Storage: dense vs TT

`tt_complexity()` reports layers without needing a fit:

```{r storage}
tab <- do.call(rbind, lapply(c(3, 5, 7, 9), function(d) {
  cx <- tt_complexity(d = d, p = 10, rank = 3)
  data.frame(
    d = d,
    npar_dense = cx$n_full,
    npar_tt = cx$n_tt_stored,
    compression = round(cx$compression_ratio, 1)
  )
}))
print(tab, row.names = FALSE)
```

At fixed \(k=10\) and \(r=3\), the dense count grows as \(10^d\); TT stored
size grows roughly linearly in \(d\). That is why GLAM / full-tensor
baselines become infeasible long before TT fits do.

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

Rank \(r\) controls **structural** capacity (and storage). It is not the
same as effective df after penalization.

# When GLAM stops

`compare_glam_tt_gaussian()` skips GLAM when \(k^d\) exceeds
`max_glam_npar` (default 5000). On higher-\(d\) grids, report that skip
honestly — do not imply TT “beat” a method that never ran.

```{r glam-skip}
# Tiny demo: raise the threshold so GLAM is skipped early
cmp <- compare_glam_tt_gaussian(
  d = 5, n_grid = 4, k = 5, ranks = 1:2,
  lambda = 1, folds = 2, max_sweeps = 3,
  max_glam_npar = 1000L,   # 5^5 = 3125 > 1000 → GLAM skipped
  seed = 1, n_test = 200,
  include_oracle = FALSE, detail_ranks = FALSE
)
cmp[, c("method", "npar", "npar_dense", "compression", "status", "reason")]
```

Full \(d=3/5\) narrative with plots: `vignette("glam-vs-tt")`.
Larger \(d=7\) scale sweeps: `compare_glam_tt_scale()` (heavier; keep out of
install-time vignette builds).

```{r scale-heavy, eval = FALSE}
# Example only — slow for package install
scale7 <- compare_glam_tt_scale(
  d = 7, n_grid_sizes = 3, k_values = c(4, 5),
  ranks = 1:2, max_sweeps = 4, max_glam_npar = 5000L
)
summarize_glam_tt_compare(scale7)
```

# Honest timings

Selecting \(r\) by \(K\)-fold CV costs many fits. Comparison helpers report
**separately**:

| Column | Meaning |
|--------|---------|
| `fit_time` | Final fit with known rank |
| `rank_selection_time` | Cost of `tt_rank_select()` |
| `total_procedure_time` | `fit_time + rank_selection_time` |
| `oracle_probe_time` | Simulation-only (not in practical totals) |

Never quote only the final `fit_time` when the paper workflow includes CV.

```{r timing-cols}
cols <- c("method", "rank", "npar", "fit_time", "rank_selection_time",
          "total_procedure_time", "status")
print(cmp[, intersect(cols, names(cmp))], row.names = FALSE)
```

# Backend: R vs Rcpp

Hot paths (ALS / PIRLS / cGCV updates) have an Rcpp backend. For production
fits prefer `backend = "auto"` (uses Rcpp when available). Vignette chunks
often force `backend = "R"` for portability.

```{r backends}
data(ishigami)
idx <- seq_len(300)
X <- as.matrix(ishigami[idx, c("x1", "x2", "x3")])
y <- ishigami$y[idx]

t_r <- system.time(
  fit_r <- ttps(
    y, X, rank = 2, k = 8, lambda = 1,
    control = tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE)
  )
)[["elapsed"]]

t_cpp <- tryCatch(
  system.time(
    fit_cpp <- ttps(
      y, X, rank = 2, k = 8, lambda = 1,
      control = tt_control(max_sweeps = 8, backend = "Rcpp", compute_edf = FALSE)
    )
  )[["elapsed"]],
  error = function(e) NA_real_
)

data.frame(
  backend = c("R", "Rcpp"),
  seconds = c(t_r, t_cpp),
  deviance = c(fit_r$deviance, if (is.finite(t_cpp)) fit_cpp$deviance else NA)
)
```

Numbers are machine-dependent; the point is the API and that Rcpp is the
fast path when compiled.

# Practical checklist

1. Report **compression** (`npar_dense` / `npar_tt`) and **EDF** separately.
2. If GLAM is skipped, say so (`status` / `reason`).
3. Split **selection time** from **final fit time**.
4. Use `backend = "auto"` or `"Rcpp"` for serious timings; keep vignette
   demos small.

# See also

- `vignette("glam-vs-tt", package = "TTPsplines")` — full-tensor vs TT demos
- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE for \(r\)
- `vignette("getting-started", package = "TTPsplines")`
- `tt_complexity()`, `compare_glam_tt_gaussian()`, `compare_glam_tt_scale()`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
