---
title: "Scalability of TT P-splines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scalability of TT P-splines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# What this vignette is

From exponential coefficient growth to linear-in-dimension TT storage.
This note separates **representational scaling**, **statistical complexity**,
and **computational cost**. It is about storage and wall time — not predictive
rank choice (`vignette("rank-selection")`) or uncertainty
(`vignette("uncertainty")`).

```{r}
library(TTPsplines)
```

# Four complexities (do not collapse them)

\[
\boxed{N_{\mathrm{full}}=\prod_j k_j}
\quad\text{(full coefficient array)}
\]

\[
\boxed{N_{\mathrm{TT}}=\sum_j r_{j-1}\,k_j\,r_j}
\quad\text{(TT storage; }r_0=r_d=1\text{)}
\]

\[
\boxed{\dim(\mathcal{M}_r)=N_{\mathrm{TT}}-\sum_{j=1}^{d-1}r_j^2}
\quad\text{(intrinsic TT manifold dimension)}
\]

\[
\boxed{\mathrm{EDF}(r,\boldsymbol\lambda)}
\quad\text{(effective statistical complexity after penalization)}
\]

Compression is not regularization, rank is not smoothness, and parameter
storage is not EDF:

\[
\boxed{r \neq \lambda \neq \mathrm{EDF}}.
\]

`tt_complexity()` reports these layers without needing a fit.

# Exact TT storage, then big-\(O\)

For uniform interior rank \(r\), equal basis size \(k_j=k\), and boundary
ranks \(r_0=r_d=1\),

\[
N_{\mathrm{TT}}=2kr+(d-2)kr^2,
\]

hence

\[
N_{\mathrm{TT}}=O(d\,k\,r^2).
\]

Examples at \(k=10\), \(r=3\): \(d=3\) gives \(60+90=150\); \(d=5\) gives
\(60+270=330\); \(d=9\) gives \(60+630=690\); \(d=10\) gives \(60+720=780\).

```{r storage}
tab <- do.call(rbind, lapply(c(3, 5, 7, 9, 10), function(d) {
  cx <- tt_complexity(d = d, p = 10, rank = 3)
  data.frame(
    d = d,
    npar_dense = cx$n_full,
    npar_tt = cx$n_tt_stored,
    compression = round(cx$compression_ratio, 1)
  )
}))
print(tab, row.names = FALSE)
```

At fixed \(k=10\) and \(r=3\), the dense count grows as \(10^d\); TT stored
size grows linearly in \(d\). At \(d=10\), \(N_{\mathrm{full}}=10^{10}\) while
\(N_{\mathrm{TT}}=780\) (compression \(\approx 1.28\times 10^{7}\)).

The coefficient dimension of conventional tensor-product representations
still grows as \(k^d\). Efficient array algorithms such as GLAM can delay the
computational consequences of this growth, but do not remove the exponential
growth of the coefficient array itself. That is why full-tensor
representations become infeasible long before TT fits do — and why a fair
comparison must say when GLAM was skipped rather than implying TT “beat” a
method that never ran.

# When GLAM stops

`compare_glam_tt_gaussian()` skips GLAM when \(k^d\) exceeds
`max_glam_npar` (default 5000). On higher-\(d\) grids, report that skip
honestly.

```{r glam-skip}
# Tiny demo: raise the threshold so GLAM is skipped early
cmp <- compare_glam_tt_gaussian(
  d = 5, n_grid = 4, k = 5, ranks = 1:2,
  lambda = 1, folds = 2, max_sweeps = 3,
  max_glam_npar = 1000L,   # 5^5 = 3125 > 1000 → GLAM skipped
  seed = 1, n_test = 200,
  include_oracle = FALSE, detail_ranks = FALSE
)
cmp[, c("method", "npar", "npar_dense", "compression", "status", "reason")]
```

Full \(d=3/5\) narrative with plots: `vignette("glam-vs-tt")`.
Larger \(d=7\) scale sweeps: `compare_glam_tt_scale()` (heavier; keep out of
install-time vignette builds).

```{r scale-heavy, eval = FALSE}
# Example only — slow for package install
scale7 <- compare_glam_tt_scale(
  d = 7, n_grid_sizes = 3, k_values = c(4, 5),
  ranks = 1:2, max_sweeps = 4, max_glam_npar = 5000L
)
summarize_glam_tt_compare(scale7)
```

For a paper-facing computational frontier, vary \(d=2,\ldots,10\) and report
full coefficient count, TT storage, implied dense memory, GLAM status, and
TT runtime. The claim is not that TT is a few times faster; it is that
**TT changes which multidimensional smooths are computationally representable**.

# Honest timings

Selecting \(r\) by \(K\)-fold CV costs many fits. Comparison helpers report
**separately**:

| Column | Meaning |
|--------|---------|
| `fit_time` | Final fit with known rank |
| `rank_selection_time` | Cost of `tt_rank_select()` |
| `total_procedure_time` | `fit_time + rank_selection_time` |
| `oracle_probe_time` | Simulation-only (not in practical totals) |

Never quote only the final `fit_time` when the paper workflow includes CV.
If the real procedure selects rank by CV,

\[
T_{\mathrm{procedure}}=T_{\mathrm{selection}}+T_{\mathrm{refit}}.
\]

A paper table should separate representation / tuning / final fit / total.

```{r timing-cols}
cols <- c("method", "rank", "npar", "fit_time", "rank_selection_time",
          "total_procedure_time", "status")
print(cmp[, intersect(cols, names(cmp))], row.names = FALSE)
```

# Backend: R vs Rcpp

Hot paths (ALS / PIRLS / cGCV updates) have an Rcpp backend. Prefer
`backend = "auto"` (uses Rcpp when available). Vignette chunks often force
`backend = "R"` for portability.

Micro-timings on tiny problems are **API demos only** — wall times of a few
milliseconds are dominated by system noise and must not be cited as speedup
evidence. Serious benchmarks should take at least tenths of a second per fit,
repeat 5–10 runs, and report medians over a grid such as
\(d\in\{3,5,7,9\}\), \(n\in\{500,2000,10000\}\), \(k\in\{8,10,15\}\).

Numerical parity (not speed) is the package validation target: same data,
bases, \(\lambda\), **initial TT cores**, convergence settings, and sweep
budget. Compare \(\eta\), \(\mu\), and deviance — not cores (TT gauge).

```{r backends}
data(ishigami)
idx <- seq_len(300)
X <- as.matrix(ishigami[idx, c("x1", "x2", "x3")])
y <- ishigami$y[idx]
init <- tt_initialize(X, rank = 2, k = 8, seed = 123)

ctrl_r <- tt_control(max_sweeps = 8, backend = "R", compute_edf = FALSE,
                     seed = 123)
fit_r <- ttps(y, X, rank = 2, k = 8, lambda = 1, init = init,
              control = ctrl_r)

fit_cpp <- tryCatch(
  ttps(
    y, X, rank = 2, k = 8, lambda = 1, init = init,
    control = tt_control(max_sweeps = 8, backend = "Rcpp",
                         compute_edf = FALSE, seed = 123)
  ),
  error = function(e) NULL
)

eta_r <- as.numeric(predict(fit_r, type = "link"))
eta_c <- if (is.null(fit_cpp)) NA_real_ else as.numeric(predict(fit_cpp, type = "link"))
data.frame(
  backend = c("R", "Rcpp"),
  deviance = c(fit_r$deviance, if (is.null(fit_cpp)) NA else fit_cpp$deviance),
  max_abs_eta_diff = c(0, if (is.null(fit_cpp)) NA else max(abs(eta_r - eta_c)))
)
```

Numbers are machine-dependent; this chunk is an API demonstration of shared
initialization, not a runtime study.

# Practical checklist

1. Report **\(N_{\mathrm{full}}\)**, **\(N_{\mathrm{TT}}\)**, **\(\dim(\mathcal{M}_r)\)**,
   and **EDF** separately.
2. Quote the exact TT identity before the \(O(dkr^2)\) statement.
3. If GLAM is skipped, say so (`status` / `reason`) — array algorithms delay
   but do not remove \(k^d\) coefficient growth.
4. Split **selection time** from **final fit time**.
5. Use `backend = "auto"` / `"Rcpp"` for serious timings; keep vignette demos
   small; validate backends by \(\eta\)/\(\mu\)/deviance parity under shared
   `init`.

# See also

- `vignette("glam-vs-tt", package = "TTPsplines")` — full-tensor vs TT demos
- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE for \(r\)
- `vignette("getting-started", package = "TTPsplines")`
- `tt_complexity()`, `compare_glam_tt_gaussian()`, `compare_glam_tt_scale()`
- Backend parity tests: `tests/testthat/test-backend-parity.R`

# References

Currie, I. D., Durban, M. and Eilers, P. H. C. (2006).
Generalized linear array models with applications to multidimensional
smoothing. *Journal of the Royal Statistical Society: Series B*, **68**,
259–280.

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
