---
title: "Uncertainty for Tensor-Train P-spline smooths"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty for Tensor-Train P-spline smooths}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

# Four distinct objects

```{r}
library(TTPsplines)
```

For a TT P-spline fit, keep these separate:

| Object | Role |
|--------|------|
| Rank \(r\) | Structural capacity of the TT coefficient factorization |
| \(\lambda\) | Directional smoothness (penalty weights) |
| EDF | Effective complexity after penalization |
| \(\mathrm{SE}(\hat f)\) | Conditional inferential uncertainty for the fitted smooth |

\[
r \neq \lambda \neq \mathrm{EDF} \neq \text{uncertainty}.
\]

v1 intervals are **approximate Bayesian / local-linearization**, **pointwise**, and
**conditional on the selected TT rank and the fitted \(\lambda\)**.
They are not simultaneous bands, and they do not include smoothing-parameter
or rank-selection uncertainty. See `docs/TT_INFERENCE_MARRA_WOOD.md`.

# Gaussian example

```{r fit}
set.seed(1)
n <- 400
X <- matrix(runif(n * 2), n, 2)
f <- sin(2 * pi * X[, 1]) * cos(2 * pi * X[, 2])
y <- f + rnorm(n, 0, 0.3)
# Unit-cube covariates → open B-spline knots on [0, 1] (not empirical
# range(X)), so prediction on seq(0,1) stays inside the knot span.
# Mild fixed λ: λ = 1 oversmooths this amplitude-1 interaction.
fit <- ttps(
  y, X, family = gaussian(), rank = 2, k = 8, lambda = 0.01,
  control = tt_control(max_sweeps = 15, seed = 1)
)
summary(fit)
```

# Pointwise SE and confidence intervals

Default covariance is Bayesian penalized-spline (\(V_B=\hat\sigma^2 H_p^{-1}\)).

```{r predict-se}
x1 <- seq(0, 1, length.out = 80)
Xslice <- cbind(x1, 0.5)
pr <- predict(
  fit, Xslice, type = "response",
  se.fit = TRUE, interval = "confidence", level = 0.95
)
head(data.frame(x1, fit = pr$fit, se = pr$se.fit,
                lower = pr$lower, upper = pr$upper))
```

Frequentist sandwich covariance is optional:

```{r vcov-types}
Vb <- vcov(fit, type = "bayesian")
Vf <- vcov(fit, type = "frequentist")
c(bayesian_trace = sum(diag(Vb)), frequentist_trace = sum(diag(Vf)))
```

`unconditional = TRUE` is not implemented (smoothing-parameter uncertainty).

# Slice plot with bands

```{r slice-plot}
truth <- sin(2 * pi * x1) * cos(2 * pi * 0.5)
plot(x1, pr$fit, type = "l", lwd = 2, ylim = range(pr$lower, pr$upper, truth),
     xlab = expression(x[1]), ylab = expression(hat(f)(x[1], 0.5)),
     main = "Pointwise 95% intervals (conditional on r, lambda)")
lines(x1, pr$lower, lty = 2, col = "grey40")
lines(x1, pr$upper, lty = 2, col = "grey40")
lines(x1, truth, col = "firebrick", lwd = 2)
legend("topright",
       c("fitted", "pointwise CI", "truth"),
       lty = c(1, 2, 1), lwd = c(2, 1, 2),
       col = c("black", "grey40", "firebrick"), bty = "n")
```

Bands are **conditional** on \((r,\lambda)\): they do not inflate for
smoothing-parameter or rank selection, and they do not correct
approximation bias. Truth outside the band can still happen under misspecified
\(\lambda\) / \(r\) or a coarse basis.

# What is still open

- GLM link-scale intervals (Gates 3–4)
- cGCV / REML smoothing-parameter uncertainty
- Rank-selection uncertainty
- Simultaneous confidence bands

# See also

- `vignette("cgcv", package = "TTPsplines")` — choosing \(\lambda\)
- `vignette("rank-selection", package = "TTPsplines")` — CV + 1-SE for \(r\)
- `vignette("aic-bic", package = "TTPsplines")` — EDF / AIC / BIC
- `vignette("getting-started", package = "TTPsplines")`
- Package note: `docs/TT_INFERENCE_MARRA_WOOD.md` (source tree)

# References

Eilers, P. H. C. and Marx, B. D. (1996).
Flexible smoothing with \(B\)-splines and penalties.
*Statistical Science*, **11**, 89–121.
<https://doi.org/10.1214/ss/1038425655>

Marra, G. and Wood, S. N. (2012).
Coverage properties of confidence intervals for generalized additive model
components. *Scandinavian Journal of Statistics*, **39**, 53–74.
<https://doi.org/10.1111/j.1467-9469.2011.00760.x>

Nychka, D. (1988).
Bayesian confidence intervals for smoothing splines.
*Journal of the American Statistical Association*, **83**, 1134–1143.
<https://doi.org/10.1080/01621459.1988.10478711>

Oseledets, I. V. (2011).
Tensor-train decomposition.
*SIAM Journal on Scientific Computing*, **33**, 2295–2317.
<https://doi.org/10.1137/090752286>

Wahba, G. (1990).
*Spline Models for Observational Data*.
SIAM, CBMS-NSF Regional Conference Series in Applied Mathematics **59**.
<https://doi.org/10.1137/1.9781611970128>

Wood, S. N. (2017).
*Generalized Additive Models: An Introduction with R* (2nd ed.).
CRC Press.
